---
title: "Cross-Tissue analysis - part 4: delta_age_trajectories"
author: "Jean-Charles de Coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
PIPELINE STAGE: 06 / Aim 2: Cross-Tissue Ageing Biology - part 4 of 10
=================================================================
Inter-tissue epigenetic age acceleration (delta-age) patterns: overall delta-age by tissue, by CpG grouping, and trends across chronological age.

Depends on objects created by part 3 (03_best_model_ranking_significance.Rmd) having already been run in the same R
session (or knitted first, with its environment preserved) - this file does not reload raw data from scratch.
=================================================================
-->

```{r config_part4}
# Set the root directory of this project (edit here, or export
# EPICLOCK_ROOT before running).
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
```

# Inter-organ delta age patterns 

A slope < 0 - epigenetic age is slowing down compared to chronological age
A R2 

```{r i}
suppressPackageStartupMessages({
library(lme4)
library(car)
})

best_predicted <- predicted_data_validation %>%
  #filter(unique_model_id %in% best$unique_model_id) %>%
  mutate(
    DeltaAge = predicted_age - Age
  )


model <- lmer(DeltaAge ~ organ + (1 | individual_fish_ID), data = best_predicted)

summary(model)
Anova(model, type = "III")
emmeans(model, pairwise ~ organ)

organ_trends <- best_predicted %>%
  group_by(organ) %>%
  do(tidy(lm(DeltaAge ~ Age, data = .))) %>%
  filter(term == "Age") %>%
  dplyr::rename(slope = estimate, p_value = p.value) %>%
  left_join(
    best_predicted %>%
      group_by(organ) %>%
      do(glance(lm(DeltaAge ~ Age, data = .))) %>%
      dplyr::select(organ, r.squared),
    by = "organ"
  )


# Ensure organ_trends is ungrouped
organ_trends <- organ_trends %>% ungroup()

n_organs <- nrow(organ_trends)


# overall delta 
p <-best_predicted %>%
  ggplot(aes(x = organ, y = DeltaAge, colour = organ)) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.15, position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point",
               size = 2, position = position_dodge(0.8)) +
  labs(
    x = "",
    y = expression(Delta~Age),
    colour = ""
  ) +
  scale_colour_manual(values = master_pal) +
  plot_theme +
  theme(legend.position = "bottom",
        axis.text.x =  element_text(angle=20, hjust =1)) 


signif_data <- get_signif_per_group(
  df = best_predicted,
  y_var = "DeltaAge",
  x_var = "organ",
  group_var = "organ"
)

# Add dynamic significance results if they exist.
if (!is.null(signif_data) && nrow(signif_data) > 0) {
  signif_data$y.position <- signif_data$y.position

  p <- p + stat_pvalue_manual(
    signif_data,
    label = "label",
    tip.length = 0.0001,
    hide.ns = FALSE,
    size = 4
  )
}

p 
delta_age_organs <- p
ggsave(paste0(output_directory,"/deltaAge_summarysummary.png"),
       delta_age_organs,
       height = 6, 
       width = 7, 
       dpi = 400)


# overall delta - by cpg_grouping

model <- lmer(DeltaAge ~ organ*cpg_grouping + (1 | individual_fish_ID), data = best_predicted)

summary(model)
Anova(model, type = "III")

p <-best_predicted %>%
  ggplot(aes(x = organ, y = DeltaAge, colour = cpg_grouping)) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.15, position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point",
               size = 2, position = position_dodge(0.8)) +
  labs(
    x = "",
    y = expression(Delta~Age),
    colour = ""
  ) +
  scale_colour_manual(values = master_pal) +

 plot_theme 



p


ggsave(paste0(output_directory,"/deltaAge_organs_summary_cpg_grouping.png"), height = 6, width = 7, dpi = 400)


# trend accross chronological time 


model <- lmer(DeltaAge ~ organ*Age + (1 | individual_fish_ID), data = best_predicted)

summary(model)
Anova(model, type = "III")
emmeans(model, pairwise ~ organ*Age)
# Assign top-right stacked positions
y_range <- range(best_predicted$DeltaAge, na.rm = TRUE)
delta   <- diff(y_range)

organ_trends <- organ_trends %>%
  arrange(desc(slope)) %>%
  mutate(
    label = sprintf("slope = %.2f, R² = %.2f", slope, r.squared),
    x_pos = max(best_predicted$Age, na.rm = TRUE) * 0.95,
    y_pos = seq(
      from = y_range[2] - 0.45 * delta,   # start 20% down from max
      to   = y_range[2] - 0.50* delta,   # end 50% down from max
      length.out = n_organs
    )
  )


organ_trends
p <-best_predicted %>%
  ggplot(aes(x = as.numeric(Age), y = DeltaAge, colour = organ, fill = organ, group = organ)) +
  geom_smooth(method = "lm", alpha = 0.2, se = TRUE, size = 0.9) +
  geom_text(
    data = organ_trends,
    aes(x = x_pos, y = y_pos, label = label, colour = organ),
    hjust = 1, size = 3.5, fontface = "italic",
    show.legend = F
  ) +
  labs(
    x = "Chronological Age",
    y = expression(Delta~Age),
    colour = "",
    fill = ""
  ) +
  scale_colour_manual(values = master_pal) +
  scale_fill_manual(values = master_pal) +

  theme(
    legend.position = "bottom"
  ) +
  plot_theme 
p 
delta_age_organs_x_age <- p
ggsave(paste0(output_directory,"/deltaAge_chro_organ_separate_models.png"),
       delta_age_organs_x_age,
       height = 6, 
       width = 5, 
       dpi = 400)

```

# ~

