---
title: "Cross-Tissue analysis - part 5: cpg_methylation_dynamics"
author: "Jean-Charles de Coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
PIPELINE STAGE: 06 / Aim 2: Cross-Tissue Ageing Biology - part 5 of 10
=================================================================
CpG-level methylation plots (mean methylation, age trajectories, sex differences) for both tissue-specific and cross-tissue-overlapping CpG sets, including heatmap statistics and cross-tissue delta-age correlation.

Depends on objects created by part 4 (04_delta_age_trajectories.Rmd) having already been run in the same R
session (or knitted first, with its environment preserved) - this file does not reload raw data from scratch.
=================================================================
-->

```{r config_part5}
# Set the root directory of this project (edit here, or export
# EPICLOCK_ROOT before running).
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
```

# CPG  Methylation plots 
### Methylation  plot levels at organ specific repeated selected CpG 

cpgs which repeatidly become selected for each organ independently 

```{r j}


# ============================================================
# Load libraries
# ============================================================
suppressPackageStartupMessages({
library(GenomicRanges)
library(dplyr)
library(purrr)
library(stringr)
})

# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
orgs <- c("Multi_organ", "Brain", "Caudal_Fin" , "Intestines")

selected_cpg <- best_selected_cpgs_organs

# ============================================================
# 2. filter methylaiton by best Cpgs 
# ============================================================


meth_list <- list()
for(o in orgs){
  cpg <- selected_cpg[[o]]
  
  
  df <- methylation_best_cpgs %>%
  filter(tile%in%  cpg)
  
  if(o !="Multi_organ"){
    df <- df %>%
      filter(Organ == o)
  }
  
   df$organ_value <- o
  
  
  meth_list[[o]] <- df
}
#============================================================
# 3. Rbind everything together into one dataframe 
#------------------------------------------------------------


merged_meth <- bind_rows(meth_list )

merged_meth <- merged_meth %>%
  mutate(organ_value = factor(organ_value, levels = c("Brain","Caudal_Fin" ,"Intestines","Multi_organ")
  ),
                        meth=methylation
  )

# Plot

p <- merged_meth %>%
    ggplot(aes(x = organ_value, y = meth, colour = organ_value, fill = organ_value, group = organ_value)) +
  ggbeeswarm::geom_quasirandom(alpha = 0.01)+
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    stat_summary(fun = mean, geom = "line", size = 0.6, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "", title = "Mean Methylation per organ") +
    scale_colour_manual(values = master_pal) +
    scale_fill_manual(values = master_pal) +
  
    theme(
        legend.position = "bottom",
    ) +
  plot_theme  +
    scale_y_continuous(
        breaks = seq(0, 100, 10),
        labels = function(x) paste0(x, "%")
    )

# Compute significance for CpG groupings dynamically (if available).
signif_data <- get_signif_per_group(
  df = merged_meth,
  y_var = "meth",
  x_var = "organ_value",
  group_var = "organ_value",
  step_value = 7
)


# Add dynamic significance results if they exist.
if (!is.null(signif_data) && nrow(signif_data) > 0) {
  signif_data$y.position <- signif_data$y.position + 5
  
  p <- p + stat_pvalue_manual(
    signif_data,
    label = "label",
    tip.length = 0.007,
    hide.ns = FALSE,
    size = 6
  )
}
p
methylation_org_specific_organs  <- p

ggsave(
  filename = paste0(output_directory, "/Methylation_cpg_best_models_organ.png"),
  plot = methylation_org_specific_organs ,
  height = 7,
  width = 6,
  dpi = 400
)

## methyaltiona cross chronological time 

p <- merged_meth %>%
    ggplot(aes(x = Age, y = meth, colour = organ_value, fill = organ_value, group = organ_value)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
  
    stat_summary(fun = mean, geom = "line", size = 0.2, position = position_dodge(0.7), colour = "#999999") +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    labs(x = "Chronological Age", y = "Mean Methylation", colour = "", fill = "") +
    scale_colour_manual(values = master_pal) +
    scale_fill_manual(values = master_pal) +
  
    theme(legend.position = "bottom"
    )  +
  plot_theme +
    scale_y_continuous(
        breaks = seq(0, 100, 10),
        labels = function(x) paste0(x, "%")
    )


p
methylation_org_specific_age_x_organs <- p

ggsave(
  filename = paste0(output_directory, "//Methylation_cpg_best_models_organ_chronological_Age.png"),
  plot = methylation_org_specific_age_x_organs,
  height = 7,
  width = 8,
  dpi = 400
)

# removing lines 
p$layers[[2]] <- NULL
p
methylation_org_specific_age_x_organs_without_line  <- p

## methyaltiona cross Sex

p <- merged_meth %>%
    ggplot(aes(x = organ_value, y = meth, colour = Sex, fill = Sex, group = Sex)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
  
    stat_summary(fun = mean, geom = "line", size = 0.2, position = position_dodge(0.7), colour = "#999999") +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "") +
    scale_colour_manual(values =master_pal) +
    scale_fill_manual(values = master_pal) +
    
  plot_theme +
    scale_y_continuous(
        breaks = seq(0, 100, 10),
        labels = function(x) paste0(x, "%")
    )+
  theme(axis.text.x = element_text(angle=20,hjust=1))


p
methylation_org_specific_sex_x_organs <- p

ggsave(
  filename = paste0(output_directory, "//Methylation_cpg_best_models_organ_Sex.png"),
  plot = methylation_org_specific_sex_x_organs,
  height = 7,
  width = 6,
  dpi = 400
)

p$layers[[2]] <- NULL
p
methylation_org_specific_sex_x_organs_without_line  <- p

## met
```

##### Statistical analysis 

```{r k}
m <- lmer( meth ~ organ_value*Age + Sex + (1 | individual_fish_ID), data = merged_meth)

summary(m)
Anova(m)
emmeans(m, pairwise ~ organ_value)
```

### Methylation  plot levels at selected cpgs selected in all organs
(Brain,intestinesmfin, Multi_organ)
cpgs which have been selected in all organs accross all models 

```{r l}


# ============================================================
# Load libraries
# ============================================================
suppressPackageStartupMessages({
library(GenomicRanges)
library(dplyr)
library(purrr)
library(stringr)
})

# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
orgs <- c("Multi_organ", "Brain", "Caudal_Fin" , "Intestines")

selected_cpg <- best_cpg_overlap_all

# ============================================================
# 2. filter methylaiton by best Cpgs 
# ============================================================


  df <- methylation_best_cpgs %>%
  filter(tile%in%  selected_cpg)
  

#============================================================
# 3. Rbind everything together into one dataframe 
#------------------------------------------------------------


merged_meth <- df

merged_meth <- merged_meth %>%
  mutate(Organ= factor(Organ, levels = c("Brain","Caudal_Fin" ,"Intestines")
  ),
                        meth=methylation
  )

# Plot

p <- merged_meth %>%
    ggplot(aes(x = Organ, y = meth, colour = Organ, fill = Organ, group = Organ)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    stat_summary(fun = mean, geom = "line", size = 0.6, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "") +
    scale_colour_manual(values = master_pal) +
    scale_fill_manual(values = master_pal) +
  
    theme(legend.position = "bottom"
    ) +
  plot_theme  +
    scale_y_continuous(
        breaks = seq(0, 100, 2),
        labels = function(x) paste0(x, "%")
    )


# Compute significance for CpG groupings dynamically (if available).
signif_data <- get_signif_per_group(
  df = merged_meth,
  y_var = "meth",
  x_var = "Organ",
  group_var = "Organ",
  step_value = 2
)


# Add dynamic significance results if they exist.
if (!is.null(signif_data) && nrow(signif_data) > 0) {
  signif_data$y.position <- signif_data$y.position 
  
  p <- p + stat_pvalue_manual(
    signif_data,
    label = "label",
    tip.length = 0.0009,
    hide.ns = FALSE,
    size = 4
  )
}
p


ggsave(
  filename = paste0(output_directory, "/Methylation_all_organ_universal_cpg.png"),
  plot = p,
  height = 5,
  width = 8,
  dpi = 400
)

## methyaltiona cross chronological time 
p <- merged_meth %>%
    ggplot(aes(x = Age, y = meth, colour = Organ, fill = Organ, group = Organ)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
  
    stat_summary(fun = mean, geom = "line", size = 0.2, position = position_dodge(0.7), colour = "#999999") +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "") +
    scale_colour_manual(values = master_pal) +
    scale_fill_manual(values = master_pal) +
  
    theme(legend.position = "bottom"
    )  +
  plot_theme +
    scale_y_continuous(
        breaks = seq(0, 100, 2),
        labels = function(x) paste0(x, "%")
    )


p

ggsave(
  filename = paste0(output_directory, "/Methylation_all_organ_universal_cpg_chronological_Age.png"),
  plot = p,
  height = 5,
  width = 8,
  dpi = 400
)


## methyaltiona cross Sex


p <- merged_meth %>%
    ggplot(aes(x = Organ, y = meth, colour = Sex, fill = Sex, group = Sex)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
  
    stat_summary(fun = mean, geom = "line", size = 0.2, position = position_dodge(0.7), colour = "#999999") +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "") +
    scale_colour_manual(values =master_pal) +
    scale_fill_manual(values = master_pal) +
    
  plot_theme +
    scale_y_continuous(
        breaks = seq(0, 100, 2),
        labels = function(x) paste0(x, "%")
    )


p


ggsave(
  filename = paste0(output_directory, "/Methylation_all_organ_universal_cpg_Sex.png"),
  plot = p,
  height = 5,
  width = 8,
  dpi = 400
)
```

#### Heatmap

```{r m}
suppressPackageStartupMessages({
  

library(pheatmap)
})

heat_df <-merged_meth %>%
  group_by(tile, Age, Organ) %>%
  summarise(
    mean_meth = mean(meth, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Age = factor(Age, levels = sort(unique(Age))),
    col_id = paste(Organ, Age, sep = "_")
  ) %>%
  dplyr::select(tile, col_id, mean_meth)


heat_mat <- heat_df %>%
  pivot_wider(names_from = col_id, values_from = mean_meth) %>%
  column_to_rownames("tile") %>%
  as.matrix()

# remove cpgs with 0 variance - will become NA in scaling - not ideal 
keep <- apply(heat_mat, 1, sd, na.rm = TRUE) > 0
heat_mat_filt <- heat_mat[keep, ]

heat_mat_scaled <- t(scale(t(heat_mat_filt)))


annotation_col <- data.frame(
  Organ = sub("_[0-9]+$", "", colnames(heat_mat_scaled)),
  Age = sub(".*_", "", colnames(heat_mat_scaled))
)

rownames(annotation_col) <- colnames(heat_mat_scaled)

# Define colors for annotations
ann_colors <- list(
  Organ = c(Brain = "#E69F00" , Caudal_Fin ="#50B562", Intestines= "#8E643B",Multi_organ ="#000000" ),  
  Age = c("3" = "#E69F00", "5" = "#8AAC94", "6" ="#3EAEC8", "8" ="#07A07D","10" = "#82C458", "12" = "#C4CF56", "14" = "#2B869D", "16" = "#606861",
          "18" = "#D4600F", "22" = "#000000") 
)
p <- tryCatch(
  {
    pheatmap(
      heat_mat_scaled,
      cluster_rows = TRUE,
      cluster_cols = TRUE,
      annotation_col = annotation_col,
      annotation_colors = ann_colors,
      show_rownames = FALSE,
      fontsize_col = 12,
      angle_col = 45,
      color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
      main = "CpG-level methylation trajectories across age and Organ"
    )
  },
  error = function(e) {
    message("Heatmap generation failed: ", conditionMessage(e))
    return(NULL)
  }
)

if (!is.null(p)) {
  ggsave(
  filename = paste0(output_directory, "/Heatmap_methylation_all_organ_universal_cpg_organ_age.png"),
  plot = p,
  height = 7,
  width = 8,
  dpi = 400
)
}



  ####
  
  
CpG_slopes <- merged_meth %>%
  mutate(Age = as.numeric(Age)) %>%
  group_by(tile, Organ) %>%
  summarise(
    slope = coef(lm(meth ~ Age))[2],
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = Organ, values_from = slope)

slope_mat <- as.matrix(CpG_slopes[, -1])
rownames(slope_mat) <- CpG_slopes$tile

p <- tryCatch(
  {
pheatmap(
  slope_mat,
  cluster_rows = TRUE,
  cluster_cols = F,
  angle_col = 45,
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
  main = "Age-associated methylation slopes per CpG"
) 

  },
  error = function(e) {
    message("Heatmap generation failed: ", conditionMessage(e))
    return(NULL)
  }
)

if (!is.null(p)) {
  ggsave(
  filename = paste0(output_directory, "/Heatmap_methylation_all_organ_universal_cpg_treatment_summary.png"),
  plot = p,
  height = 13,
  width = 8,
  dpi = 400
)
}
  


  
  # viewing top tiles 
  # Compute slope difference
CpG_diff <- CpG_slopes %>%
  mutate(
    slope_diff = Brain - Caudal_Fin - Intestines  # difference in age-associated slope
  ) %>%
  arrange(desc(abs(slope_diff)))  # rank by absolute difference

# View top divergent tiles
top_CpGs <- CpG_diff %>%
  slice_head(n = 20)  # top 20 tiles with largest treatment effect

top_CpGs
```

##### Statistical analysis 

```{r m1}
m <- lmer( meth ~ Organ*Age + Sex + (1 | individual_fish_ID), data = merged_meth)

summary(m)
Anova(m)
emmeans(m, pairwise ~ Organ)
```


### Methylation  plot levels at selected cpgs selected in all singletissue  somatic organs
( brain, intestines, flin clip  )

cpgs which have been selected in all single tissue organs accross all models 
```{r n}


# ============================================================
# Load libraries
# ============================================================
suppressPackageStartupMessages({
library(GenomicRanges)
library(dplyr)
library(purrr)
library(stringr)
})


# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
orgs <- c( "Brain", "Caudal_Fin" , "Intestines")

selected_cpg <- best_cpg_overlap_single

# ============================================================
# 2. filter methylaiton by best Cpgs 
# ============================================================

  df <- methylation_best_cpgs %>%
  filter(tile%in%  selected_cpg)
  

#============================================================
# 3. Rbind everything together into one dataframe 
#------------------------------------------------------------


merged_meth <- df

merged_meth <- merged_meth %>%
  mutate(Organ = factor(Organ, levels = c("Brain","Caudal_Fin" ,"Intestines","Multi_organ")
  ),
                        meth=methylation
  )


# Plot

p <- merged_meth %>%
    ggplot(aes(x = Organ, y = meth, colour = Organ, fill = Organ, group = Organ)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    stat_summary(fun = mean, geom = "line", size = 0.6, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "", title = "Mean Methylation per organ") +
    scale_colour_manual(values = master_pal) +
    scale_fill_manual(values = master_pal) +
  
    theme(legend.position = "bottom"
    ) +
  plot_theme  +
    scale_y_continuous(
        breaks = seq(0, 100, 2),
        labels = function(x) paste0(x, "%")
    )


# Compute significance for CpG groupings dynamically (if available).
signif_data <- get_signif_per_group(
  df = merged_meth,
  y_var = "meth",
  x_var = "Organ",
  group_var = "Organ",
  step_value = 2
)


# Add dynamic significance results if they exist.
if (!is.null(signif_data) && nrow(signif_data) > 0) {
  signif_data$y.position <- signif_data$y.position 
  
  p <- p + stat_pvalue_manual(
    signif_data,
    label = "label",
    tip.length = 0.001,
    hide.ns = FALSE,
    size = 4
  )
}
p

methylation_overlap_soma <- p

ggsave(
  filename = paste0(output_directory, "/Methylation_single_organ_universal_cpg.png"),
  plot = methylation_overlap_soma,
  height = 7,
  width = 6,
  dpi = 400
)

## methyaltiona cross chronological time 

p <- merged_meth %>%
    ggplot(aes(x = Age, y = meth, colour = Organ, fill = Organ, group = Organ)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
  
    stat_summary(fun = mean, geom = "line", size = 0.2, position = position_dodge(0.7), colour = "#999999") +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    labs( y = "Mean Methylation", colour = "", fill = "", x = "Chronological Age") +
    scale_colour_manual(values = master_pal) +
    scale_fill_manual(values = master_pal) +
  
    theme(legend.position = "bottom"
    )  +
  plot_theme +
    scale_y_continuous(
        breaks = seq(0, 100, 5),
        labels = function(x) paste0(x, "%")
    )


p
methylation_overlap_soma_age_x_organs <- p

ggsave(
  filename = paste0(output_directory, "/Methylation_single_organ_universal_cpg_chronological_Age.png"),
  plot = methylation_overlap_soma_age_x_organs,
  height = 5,
  width = 8,
  dpi = 400
)

# removing Lines form plot 
p$layers[[2]] <- NULL
p
methylation_overlap_soma_age_x_organs_without_line  <- p

## methyaltiona cross Sex

p <- merged_meth %>%
    ggplot(aes(x = Organ, y = meth, colour = Sex, fill = Sex, group = Sex)) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.6,
                 alpha = 0.5, position = position_dodge(0.7)) +
  
    stat_summary(fun = mean, geom = "line", size = 0.2, position = position_dodge(0.7), colour = "#999999") +
    stat_summary(fun = mean, geom = "point", size = 2, position = position_dodge(0.7)) +
    labs(x = "", y = "Mean Methylation", colour = "", fill = "") +
    scale_colour_manual(values =master_pal) +
    scale_fill_manual(values = master_pal) +
    
  plot_theme +
    scale_y_continuous(
        breaks = seq(0, 100, 2),
        labels = function(x) paste0(x, "%")
    ) +
  theme(axis.text.x = element_text(angle=20, hjust=1))+
  theme(legend.position = "bottom")


p
methylation_overlap_soma_sex_x_organs <- p

ggsave(
  filename = paste0(output_directory, "/Methylation_single_organ_universal_cpg_Sex.png"),
  plot = methylation_overlap_soma_sex_x_organs,
  height = 6,
  width = 7,
  dpi = 400
)

# removing Lines form plot 
p$layers[[2]] <- NULL
p
methylation_overlap_soma_sex_x_organs_without_line  <- p

```
#### Heatmap

```{r o}
suppressPackageStartupMessages({
  

library(pheatmap)
})

heat_df <-merged_meth %>%
  group_by(tile, Age, Organ) %>%
  summarise(
    mean_meth = mean(meth, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Age = factor(Age, levels = sort(unique(Age))),
    col_id = paste(Organ, Age, sep = "_")
  ) %>%
  dplyr::select(tile, col_id, mean_meth)


heat_mat <- heat_df %>%
  pivot_wider(names_from = col_id, values_from = mean_meth) %>%
  column_to_rownames("tile") %>%
  as.matrix()

# remove cpgs with 0 variance - will become NA in scaling - not ideal 
keep <- apply(heat_mat, 1, sd, na.rm = TRUE) > 0
heat_mat_filt <- heat_mat[keep, ]

heat_mat_scaled <- t(scale(t(heat_mat)))


annotation_col <- data.frame(
  Organ = sub("_[0-9]+$", "", colnames(heat_mat_scaled)),
  Age = sub(".*_", "", colnames(heat_mat_scaled))
)

rownames(annotation_col) <- colnames(heat_mat_scaled)

# Define colors for annotations
ann_colors <- list(
  Organ = c(Brain = "#E69F00" , Caudal_Fin ="#50B562", Intestines= "#8E643B"),  
  Age = c("3" = "#E69F00", "5" = "#8AAC94", "6" ="#3EAEC8", "8" ="#07A07D","10" = "#82C458", "12" = "#C4CF56", "14" = "#2B869D", "16" = "#606861",
          "18" = "#D4600F", "22" = "#000000") 
)
p<- pheatmap(
  heat_mat_scaled,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  show_rownames = FALSE,
  fontsize_col = 12,
  angle_col = 45,           # <- rotates column labels
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
  main = "CpG-level methylation trajectories across age and Organ"
)
p 
 heatmap_overlap_soma  <- p

 ggsave(
  filename = paste0(output_directory, "/Heatmap_methylation_single_organ_universal_cpg_organ_age.png"),
  plot =  heatmap_overlap_soma ,
  height = 7,
  width = 8,
  dpi = 400
)
  ####
  
  
CpG_slopes <- merged_meth %>%
  mutate(Age = as.numeric(Age)) %>%
  group_by(tile, Organ) %>%
  summarise(
    slope = coef(lm(meth ~ Age))[2],
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = Organ, values_from = slope)

slope_mat <- as.matrix(CpG_slopes[, -1])
rownames(slope_mat) <- CpG_slopes$tile

p<- pheatmap(
  slope_mat,
  cluster_rows = TRUE,
  cluster_cols = F,
  angle_col = 45,
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
  main = "Age-associated methylation slopes per CpG"
) 

  p 
  summary_heatmap_overlap_soma  <-p 
  
  ggsave(
  filename = paste0(output_directory, "/Heatmap_methylation_single_organ_universal_cpg_treatment_summary.png"),
  plot =  summary_heatmap_overlap_soma ,
  height = 8,
  width = 6,
  dpi = 400
)

  
  # viewing top tiles 
  # Compute slope difference
CpG_diff <- CpG_slopes %>%
  mutate(
    slope_diff = Brain - Caudal_Fin - Intestines  # difference in age-associated slope
  ) %>%
  arrange(desc(abs(slope_diff)))  # rank by absolute difference

# View top divergent tiles
top_CpGs <- CpG_diff %>%
  slice_head(n = 20)  # top 20 tiles with largest treatment effect

top_CpGs
```

