---
title: "Cross-Tissue analysis - part 6: cpg_functional_annotation_enrichment"
author: "Jean-Charles de Coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
PIPELINE STAGE: 06 / Aim 2: Cross-Tissue Ageing Biology - part 6 of 10
=================================================================
Genomic context and GO/KEGG functional enrichment of organ-specific and universal (all-tissue, all-somatic, all-single-tissue) overlapping CpGs; genomic distribution plots of selected CpGs across three CpG sets.

Depends on objects created by part 5 (05_cpg_methylation_dynamics.Rmd) having already been run in the same R
session (or knitted first, with its environment preserved) - this file does not reload raw data from scratch.
=================================================================
-->

```{r config_part6}
# Set the root directory of this project (edit here, or export
# EPICLOCK_ROOT before running).
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
```

##### Statistical analysis 

```{r o1}

suppressPackageStartupMessages({
  library(lme4)
  library(car)
  library(emmeans)
  library(vegan)
  library(cluster)
}
)
# overall 
m <- lmer( meth ~ Organ*Age + Sex + (1 | individual_fish_ID), data = merged_meth)

summary(m)
Anova(m)

## Heatmap statistics
## ------------------------------------------------------------
## 1. Metadata preparation (must match heatmap input)
## ------------------------------------------------------------

meta <- annotation_col %>%
  rownames_to_column("sample") %>%
  as_tibble()

stopifnot(all(meta$sample == colnames(heat_mat_scaled)))

## ------------------------------------------------------------
## 2. CpG-level age trajectory correlations between organs
## ------------------------------------------------------------

long_df <- heat_mat_scaled %>%
  as.data.frame() %>%
  rownames_to_column("CpG") %>%
  pivot_longer(-CpG, names_to = "sample", values_to = "meth") %>%
  left_join(meta, by = "sample")

traj_cor <- long_df %>%
  mutate(Age = as.numeric(Age)) %>%
  select(CpG, Age, Organ, meth) %>%
  pivot_wider(
    names_from = Organ,
    values_from = meth
  ) %>%
  group_by(CpG) %>%
  summarise(
    cor_brain_intestine = cor(
      Brain,
      Intestines,
      use = "pairwise.complete.obs"
    ),
    cor_brain_fin = cor(
      Brain,
      Caudal_Fin,
      use = "pairwise.complete.obs"
    ),
    cor_intestine_fin = cor(
      Intestines,
      Caudal_Fin,
      use = "pairwise.complete.obs"
    ),
    .groups = "drop"
  )

trajectory_summary <- traj_cor %>%
  summarise(
    median_brain_intestine = median(cor_brain_intestine, na.rm = TRUE),
    median_brain_fin       = median(cor_brain_fin, na.rm = TRUE),
    median_intestine_fin   = median(cor_intestine_fin, na.rm = TRUE)
  )

print(trajectory_summary)


## ------------------------------------------------------------
## 3. Multivariate test of tissue separation (PERMANOVA)
## ------------------------------------------------------------

dist_mat <- dist(t(heat_mat_scaled), method = "euclidean")

adonis_res <- adonis2(
  dist_mat ~ Organ,
  data = meta,
  permutations = 999
)

print(adonis_res)



```

# ~ 


## Cross-tissue correlation and ΔAge discordance
 Multi-tissue and cor-tissue comparaison 
Variance in residulas ( delta age) was assessed accross tissues to quantify tissue specific regulation. Large divergences point to metabolic or proliferative differences -  https://doi.org/10.7554/eLife.40675

I dentified "fast-ageing" adn "slow-ageing" organs within individuals to link tissue-specific maintenance or regeneration capability -https://doi.org/10.18632/aging.206182


Compare cross-tissue correlations and predicted DNAm - for each individual, predicted age accross tissues was correlated - exposing systemic synchrony vers organ specialisation  - https://doi.org/10.18632/aging.206182

This tests systemic vs. organ-specific synchrony.
```{r p}


# Combine predicted ages


pred_df <- predicted_data_validation 

# 1 Compute ΔAge (age acceleration residuals) - loads of Na's so had to remove them 
pred_df <- pred_df %>%
  #filter(organ != "Multi_organ")%>%
  group_by(organ) %>%
  group_modify(~ {
    df <- .x
    complete <- complete.cases(df$predicted_age, df$Age)
    df$delta_age <- NA_real_
    df$delta_age[complete] <- resid(lm(predicted_age ~Age, data = df[complete, ]))
    df
  }) %>%
  ungroup()

# if any individual id's missing compute them 
#pred_df <- pred_df %>% mutate(individual_fish_ID = as.numeric(sub("[A-Za-z]+$", "", original_tube_ID)))
missing <- pred_df$individual_fish_ID == "" | grepl("^\\s*$", pred_df$individual_fish_ID)
pred_df$individual_fish_ID[missing] <-
  gsub("[A-Za-z]", "", pred_df$original_tube_ID[missing])


pred_df <- pred_df %>% mutate(unique_id = paste(pred_df$individual_fish_ID,cpg_grouping, sep = "_"))

# 2️  check correlations of ΔAge across organs (requires wide format) - Also required unique individual ID - we want to see how does a single individual organs differ 
delta_wide <- pred_df %>%
  filter(organ != "Multi_organ") %>%
  dplyr::select(sample_id = individual_fish_ID , Organ, delta_age) %>%
  tidyr::pivot_wider(names_from = Organ, values_from = delta_age)

delta_summarised <- pred_df %>%
  filter(organ != "Multi_organ") %>%
  group_by(sample_id = individual_fish_ID, organ) %>%
  summarise(delta_age = mean(delta_age), .groups = "drop")%>%
  pivot_wider(names_from = organ, values_from = delta_age)


# Correlation matrix between organs
delta_cor <- cor(delta_summarised[,-1], use = "pairwise.complete.obs")
print(round(delta_cor, 2)) # the closer the correlation is to 1  the stronger they are related so if Brain-intestine is -0.69 - both are more likely to age together - is over 0.7 it is interesting 
# here Brain - fin = 1, fin-intestine = -0.83 INTERESTING

#pairs(delta_wide[,-1], main = "ΔAge correlations between organs")

# 3 Plot ΔAge per individual to see organ offsets
p<- ggplot(pred_df, aes(x = organ, y = delta_age, colour = organ, fill = organ)) +
  introdataviz::geom_split_violin(alpha = 0.2, show.legend = F)+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, size = 0.7) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  # stat_summary(fun = mean, geom = "line", group = 1, linetype = "dashed") +
  scale_colour_manual(values = master_pal) +
  scale_fill_manual(values = master_pal) +
  labs(
    y = expression(Delta*"Age"),
    x = "",
    colour = "",
    title = "Delta-Age correlation"
  ) +
  guides(fill = "none")+
  plot_theme +

   theme(    
    legend.position = "bottom",
    axis.text.x = element_text(angle=20,hjust=1)
  ) 
p
cross_tissue_delta <- p

ggsave(
  filename = paste0(output_directory, "/ΔAge_organ_best_model.png"),
  plot = cross_tissue_delta,
  height = 6,
  width = 7,
  dpi = 400
)
```


### ~
# Function and genomic context of CpGs

 Functional enrichment of CpGs driving each organ clock

Function enrichment of genes near clock CpGs - GO term, KEGG, Chromatin state, enhancer enrichment - Functionally enriched CpGs often cluster in genes related to senesccne, Polycomb targets (PCGTs) and proliferation control - https://doi.org/10.1038/s43856-025-00739-4

missMethyl is used to account for probe bias.

## Organ specific reoccuring cpgs 
```{r q, echo=T, message=FALSE, warning=FALSE, results='hide'}
###############################################################
# Epigenetic Clock CpG Functional Annotation and Enrichment
# Purpose: Identify functional enrichment (GO / KEGG) of CpGs 
#          contributing to best-performing organ-specific 
#          epigenetic clock models in Danio rerio.
###############################################################

#--------------------------------------------------------------
# 0. Load Required Packages
#--------------------------------------------------------------
suppressPackageStartupMessages({
library(missMethyl)             # CpG-centric enrichment (gometh)
library(dplyr)                  # Data manipulation
library(GenomicRanges)          # Genomic interval operations
library(TxDb.Drerio.UCSC.danRer11.refGene) # Zebrafish gene annotation
library(clusterProfiler)        # Classical GO/KEGG enrichment
library(org.Dr.eg.db)           # Zebrafish gene ontology database
library(gprofiler2)             # Orthogonal enrichment tool (g:Profiler)
})
#--------------------------------------------------------------
# 1. Condense Model Metrics: Select the Best Model per Organ
#--------------------------------------------------------------
# 'merged_data$metrics' contains model performance metrics 
# (e.g., mean absolute error [MAE] and R² from training).



#--------------------------------------------------------------
# 2. Rank Models per Organ by Combined Performance
#--------------------------------------------------------------
# A composite ranking is computed as:
# - lower MAE = better
# - higher R² = better
# - combined rank = mae_rank + r2_rank
# The top-performing model per organ is retained.

#--------------------------------------------------------------
# 3. Extract CpGs from the Best-Performing Models
#--------------------------------------------------------------
# 'cpg_data_validation' contains model-specific CpGs; extract those
# corresponding to the top-ranked models per organ.

Multi_organ_best_cpg <- tibble::tibble(
  organ = names(best_selected_cpgs_organs),
  cpgs  = best_selected_cpgs_organs
) |>
  tidyr::unnest_longer(cpgs, values_to = "cpg")

Multi_organ_best_cpg  <- Multi_organ_best_cpg %>%
  pull(unique(cpg))

#--------------------------------------------------------------
# 5. Extract Candidate and Selected CpGs
#--------------------------------------------------------------
# Candidate CpGs = all CpGs available for model training.
# Selected CpGs = CpGs retained in the final best model.
# Both are represented as genomic coordinate strings 
# (e.g., "chr1:24557370:24557869").

Multi_organ_candidate_cpg <- methylation_data_list$Multi_organ$Probe
Brain_candidate_cpg     <- methylation_data_list$Brain$Probe
Caudal_Fin_candidate_cpg       <- methylation_data_list$Caudal_Fin$Probe
Intestines_candidate_cpg <- methylation_data_list$Intestines$Probe

Multi_organ_selected_cpg  <- best_selected_cpgs_organs$Multi_organ
Brain_selected_cpg      <-best_selected_cpgs_organs$Brain
Caudal_Fin_selected_cpg        <-best_selected_cpgs_organs$Caudal_Fin
Intestines_selected_cpg  <-best_selected_cpgs_organs$Intestines


# need to make sure cpgs match annotation_database 

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}

Multi_organ_candidate_cpg <-adding_chr_cpg(Multi_organ_candidate_cpg)
Brain_candidate_cpg     <-adding_chr_cpg(Brain_candidate_cpg  )
Caudal_Fin_candidate_cpg       <-adding_chr_cpg(Caudal_Fin_candidate_cpg   )
Intestines_candidate_cpg <-adding_chr_cpg(Intestines_candidate_cpg)

Multi_organ_selected_cpg  <- adding_chr_cpg(Multi_organ_selected_cpg)
Brain_selected_cpg      <-adding_chr_cpg(Brain_selected_cpg )
Caudal_Fin_selected_cpg <- adding_chr_cpg(Caudal_Fin_selected_cpg)
Intestines_selected_cpg  <-adding_chr_cpg(Intestines_selected_cpg)

#--------------------------------------------------------------
# 6. Functional Annotation Loop across Organs
#--------------------------------------------------------------
# For each organ, CpGs are:
#  (i) parsed into GRanges,
#  (ii) mapped to the nearest genes using TxDb,
#  (iii) subjected to multiple enrichment analyses:
#       - missMethyl::gometh (bias-corrected CpG enrichment)
#       - clusterProfiler::enrichGO (classical GO)
#       - gprofiler2::gost (independent validation)
# Results are stored in a nested list (organ_list).

organ_list <- list()
organs <- c( "Multi_organ","Brain", "Caudal_Fin" , "Intestines")

for (organ in organs) {
  message("\n--- Processing ", organ, " ---")

  organ_list[[organ]] <- list()

  tryCatch({

    #----------------------------------------------------------
    # 6.1 Retrieve CpG Vectors
    #----------------------------------------------------------
    if (!exists(paste0(organ, "_selected_cpg"))) stop("Missing selected CpG object")
    if (!exists(paste0(organ, "_candidate_cpg"))) stop("Missing candidate CpG object")

    selected_cpg <- get(paste0(organ, "_selected_cpg"))
    candidate_cpg <- get(paste0(organ, "_candidate_cpg"))

    message("Selected CpGs: ", length(selected_cpg),
            " | Candidate CpGs: ", length(candidate_cpg))

    #----------------------------------------------------------
    # 6.2 Parse Genomic Coordinates
    #----------------------------------------------------------
    # Each CpG name is split into chromosome, start, and end.
    selected_cpg_df <- data.frame(selected_cpgs = selected_cpg, stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpgs, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

    candidate_cpg_df <- data.frame(selected_cpgs = candidate_cpg, stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpgs, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

    #----------------------------------------------------------
    # 6.3 Construct GRanges Objects
    #----------------------------------------------------------
    selected_gr <- GRanges(seqnames = selected_cpg_df$chr,
                           ranges = IRanges(start = selected_cpg_df$start,
                                            end = selected_cpg_df$end))

    candidate_gr <- GRanges(seqnames = candidate_cpg_df$chr,
                            ranges = IRanges(start = candidate_cpg_df$start,
                                             end = candidate_cpg_df$end))

    #----------------------------------------------------------
    # 6.4 Map CpGs to the Nearest Genes
    #----------------------------------------------------------
    txdb <- TxDb.Drerio.UCSC.danRer11.refGene
    genes_gr <- genes(txdb)

    selected_entrez  <- names(genes_gr)[nearest(selected_gr, genes_gr)]
    candidate_entrez <- names(genes_gr)[nearest(candidate_gr, genes_gr)]

    # Store gene sets for manual inspection
    organ_list[[organ]]$selected_genes  <- selected_entrez
    organ_list[[organ]]$candidate_genes <- candidate_entrez

    message("Mapped ", length(selected_entrez), " selected CpGs to genes.")
    message("Mapped ", length(candidate_entrez), " candidate CpGs to genes.")

    #----------------------------------------------------------
    # 6.5 CpG-Level Enrichment using missMethyl (GO + KEGG)
    #----------------------------------------------------------
    organ_list[[organ]]$gst <- tryCatch({
      gometh(sig.cpg = selected_entrez,
             all.cpg = candidate_entrez,
             collection = "GO",
             plot.bias = TRUE)
    }, error = function(e) {
      message("gometh GO failed: ", e$message)
      NULL
    })

    organ_list[[organ]]$gst_kegg <- tryCatch({
      gometh(sig.cpg = selected_entrez,
             all.cpg = candidate_entrez,
             collection = "KEGG",
             plot.bias = TRUE)
    }, error = function(e) {
      message("gometh KEGG failed: ", e$message)
      NULL
    })

    #----------------------------------------------------------
    # 6.6 Classical GO Enrichment via clusterProfiler
    #----------------------------------------------------------
    message("Starting - GO enrichment")

    selected_genes  <- unique(selected_entrez)
    candidate_genes <- unique(candidate_entrez)

    organ_list[[organ]]$ego <- enrichGO(
      gene          = selected_genes,
      universe      = candidate_genes,
      OrgDb         = org.Dr.eg.db,
      keyType       = "ENTREZID",
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.2,
      readable      = TRUE
    )

    # Generate a dotplot if results exist
    if (nrow(as.data.frame(organ_list[[organ]]$ego)) > 0) {
      organ_list[[organ]]$enrich_dotplot <- dotplot(
        organ_list[[organ]]$ego,
        showCategory = 20
      )
    } else {
      message("No GO enrichment terms found for ", organ)
    }

    #----------------------------------------------------------
    # 6.7 Functional Enrichment with g:Profiler
    #----------------------------------------------------------
    warning("Starting - g:Profiler enrichment")

    organ_list[[organ]]$gp <- gost(
      query              = selected_genes,
      organism           = "drerio",
      correction_method  = "fdr",
      sources            = c("GO:BP", "GO:MF", "GO:CC", "KEGG")
    )

    # Visualise g:Profiler results if available
if (!is.null(organ_list[[organ]]$gp) && !is.null(organ_list[[organ]]$gp$result)) {

  # Store the result table separately for convenience
  gp_res <- organ_list[[organ]]$gp$result

  # Limit to top 20 enriched terms by adjusted p-value
  top_terms <- gp_res %>%
    arrange(p_value) 
   # dplyr::slice_head(n = 20)

  # Define colour scale based on number of unique cpg_groupings
  n_sources <- length(unique(top_terms$source))
  fill_cols <- okabe_ito_palette(max(3, n_sources))  # ensure enough distinct colours

  # Generate enrichment plot
  organ_list[[organ]]$gp_plot <- ggplot(top_terms, aes(
      x = reorder(term_name, -log10(p_value)),
      y = -log10(p_value),
      fill = source
    )) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal(base_size = 12) +
    scale_fill_manual(values = fill_cols, name = "Database") +
    labs(
      title = paste("g:Profiler enrichment -", organ),
      x = NULL,
      y = expression(-log[10](adjusted~p~value))
    ) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    ) +
  plot_theme 

  print(organ_list[[organ]]$gp_plot)
  
  ggsave(
  filename = paste0(output_directory, "/Functional_enrichment_organ_specific_GOterm_",organ,".png"),
  plot = organ_list[[organ]]$gp_plot,
  height = 6,
  width = 10,
  dpi = 400
)
  
  enriched_terms <- as.data.frame(organ_list[[organ]][["gp"]][["result"]])
enriched_terms$term_name

write_csv(enriched_terms,paste0("results/Profile_enrichment_organ_specific_",organ,".csv"))
}

  }, error = function(e) {
    message("Processing failed for ", organ, ": ", e$message)
  })
}



```




## Universal all organs overlapping cpgs 
```{r r, echo=T, message=FALSE, warning=FALSE, results='hide'}
###############################################################
# Epigenetic Clock CpG Functional Annotation and Enrichment
# Purpose: Identify functional enrichment (GO / KEGG) of CpGs 
#          contributing to best-performing organ-specific 
#          epigenetic clock models in Danio rerio.
###############################################################

#--------------------------------------------------------------
# 0. Load Required Packages
#--------------------------------------------------------------
suppressPackageStartupMessages({
library(missMethyl)             # CpG-centric enrichment (gometh)
library(dplyr)                  # Data manipulation
library(GenomicRanges)          # Genomic interval operations
library(TxDb.Drerio.UCSC.danRer11.refGene) # Zebrafish gene annotation
library(clusterProfiler)        # Classical GO/KEGG enrichment
library(org.Dr.eg.db)           # Zebrafish gene ontology database
library(gprofiler2)             # Orthogonal enrichment tool (g:Profiler)
})
#--------------------------------------------------------------
# 1. Condense Model Metrics: Select the Best Model per Organ
#--------------------------------------------------------------
# 'merged_data$metrics' contains model performance metrics 
# (e.g., mean absolute error [MAE] and R² from training).



#--------------------------------------------------------------
# 2. Rank Models per Organ by Combined Performance
#--------------------------------------------------------------
# A composite ranking is computed as:
# - lower MAE = better
# - higher R² = better
# - combined rank = mae_rank + r2_rank
# The top-performing model per organ is retained.

#--------------------------------------------------------------
# 3. Extract CpGs from the Best-Performing Models
#--------------------------------------------------------------
# 'cpg_data_validation' contains model-specific CpGs; extract those
# corresponding to the top-ranked models per organ.

Multi_organ_best_cpg  <- unique(best_cpg_overlap_all)

#--------------------------------------------------------------
# 5. Extract Candidate and Selected CpGs
#--------------------------------------------------------------
# Candidate CpGs = all CpGs available for model training.
# Selected CpGs = CpGs retained in the final best model.
# Both are represented as genomic coordinate strings 
# (e.g., "chr1:24557370:24557869").

Multi_organ_candidate_cpg <- methylation_data_list$Multi_organ$Probe
Brain_candidate_cpg     <- methylation_data_list$Brain$Probe
Caudal_Fin_candidate_cpg       <- methylation_data_list$Caudal_Fin$Probe
Intestines_candidate_cpg <- methylation_data_list$Intestines$Probe

Multi_organ_selected_cpg  <- Multi_organ_best_cpg 
Brain_selected_cpg      <-Multi_organ_best_cpg 
Caudal_Fin_selected_cpg    <- Multi_organ_best_cpg 
Intestines_selected_cpg  <-Multi_organ_best_cpg 


# need to make sure cpgs match annotation_database 

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}

Multi_organ_candidate_cpg <-adding_chr_cpg(Multi_organ_candidate_cpg)
Brain_candidate_cpg     <-adding_chr_cpg(Brain_candidate_cpg  )
Caudal_Fin_candidate_cpg       <-adding_chr_cpg(Caudal_Fin_candidate_cpg   )
Intestines_candidate_cpg <-adding_chr_cpg(Intestines_candidate_cpg)

Multi_organ_selected_cpg  <- adding_chr_cpg(Multi_organ_selected_cpg)
Brain_selected_cpg      <-adding_chr_cpg(Brain_selected_cpg )
Caudal_Fin_selected_cpg <- adding_chr_cpg(Caudal_Fin_selected_cpg)
Intestines_selected_cpg  <-adding_chr_cpg(Intestines_selected_cpg)

#--------------------------------------------------------------
# 6. Functional Annotation Loop across Organs
#--------------------------------------------------------------
# For each organ, CpGs are:
#  (i) parsed into GRanges,
#  (ii) mapped to the nearest genes using TxDb,
#  (iii) subjected to multiple enrichment analyses:
#       - missMethyl::gometh (bias-corrected CpG enrichment)
#       - clusterProfiler::enrichGO (classical GO)
#       - gprofiler2::gost (independent validation)
# Results are stored in a nested list (organ_list).

organ_list <- list()
organs <- c( "Multi_organ","Brain", "Caudal_Fin" , "Intestines")

for (organ in organs) {
  message("\n--- Processing ", organ, " ---")

  organ_list[[organ]] <- list()

  tryCatch({

    #----------------------------------------------------------
    # 6.1 Retrieve CpG Vectors
    #----------------------------------------------------------
    if (!exists(paste0(organ, "_selected_cpg"))) stop("Missing selected CpG object")
    if (!exists(paste0(organ, "_candidate_cpg"))) stop("Missing candidate CpG object")

    selected_cpg <- get(paste0(organ, "_selected_cpg"))
    candidate_cpg <- get(paste0(organ, "_candidate_cpg"))

    message("Selected CpGs: ", length(selected_cpg),
            " | Candidate CpGs: ", length(candidate_cpg))

    #----------------------------------------------------------
    # 6.2 Parse Genomic Coordinates
    #----------------------------------------------------------
    # Each CpG name is split into chromosome, start, and end.
    selected_cpg_df <- data.frame(selected_cpgs = selected_cpg, stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpgs, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

    candidate_cpg_df <- data.frame(selected_cpgs = candidate_cpg, stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpgs, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

    #----------------------------------------------------------
    # 6.3 Construct GRanges Objects
    #----------------------------------------------------------
    selected_gr <- GRanges(seqnames = selected_cpg_df$chr,
                           ranges = IRanges(start = selected_cpg_df$start,
                                            end = selected_cpg_df$end))

    candidate_gr <- GRanges(seqnames = candidate_cpg_df$chr,
                            ranges = IRanges(start = candidate_cpg_df$start,
                                             end = candidate_cpg_df$end))

    #----------------------------------------------------------
    # 6.4 Map CpGs to the Nearest Genes
    #----------------------------------------------------------
    txdb <- TxDb.Drerio.UCSC.danRer11.refGene
    genes_gr <- genes(txdb)

    selected_entrez  <- names(genes_gr)[nearest(selected_gr, genes_gr)]
    candidate_entrez <- names(genes_gr)[nearest(candidate_gr, genes_gr)]

    # Store gene sets for manual inspection
    organ_list[[organ]]$selected_genes  <- selected_entrez
    organ_list[[organ]]$candidate_genes <- candidate_entrez

    message("Mapped ", length(selected_entrez), " selected CpGs to genes.")
    message("Mapped ", length(candidate_entrez), " candidate CpGs to genes.")

    #----------------------------------------------------------
    # 6.5 CpG-Level Enrichment using missMethyl (GO + KEGG)
    #----------------------------------------------------------
    organ_list[[organ]]$gst <- tryCatch({
      gometh(sig.cpg = selected_entrez,
             all.cpg = candidate_entrez,
             collection = "GO",
             plot.bias = TRUE)
    }, error = function(e) {
      message("gometh GO failed: ", e$message)
      NULL
    })

    organ_list[[organ]]$gst_kegg <- tryCatch({
      gometh(sig.cpg = selected_entrez,
             all.cpg = candidate_entrez,
             collection = "KEGG",
             plot.bias = TRUE)
    }, error = function(e) {
      message("gometh KEGG failed: ", e$message)
      NULL
    })

    #----------------------------------------------------------
    # 6.6 Classical GO Enrichment via clusterProfiler
    #----------------------------------------------------------
    message("Starting - GO enrichment")

    selected_genes  <- unique(selected_entrez)
    candidate_genes <- unique(candidate_entrez)

    organ_list[[organ]]$ego <- enrichGO(
      gene          = selected_genes,
      universe      = candidate_genes,
      OrgDb         = org.Dr.eg.db,
      keyType       = "ENTREZID",
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.2,
      readable      = TRUE
    )

    # Generate a dotplot if results exist
    if (nrow(as.data.frame(organ_list[[organ]]$ego)) > 0) {
      organ_list[[organ]]$enrich_dotplot <- dotplot(
        organ_list[[organ]]$ego,
        showCategory = 20
      )
    } else {
      message("No GO enrichment terms found for ", organ)
    }

    #----------------------------------------------------------
    # 6.7 Functional Enrichment with g:Profiler
    #----------------------------------------------------------
    warning("Starting - g:Profiler enrichment")

    organ_list[[organ]]$gp <- gost(
      query              = selected_genes,
      organism           = "drerio",
      correction_method  = "fdr",
      sources            = c("GO:BP", "GO:MF", "GO:CC", "KEGG")
    )

    # Visualise g:Profiler results if available
if (!is.null(organ_list[[organ]]$gp) && !is.null(organ_list[[organ]]$gp$result)) {

  # Store the result table separately for convenience
  gp_res <- organ_list[[organ]]$gp$result

  # Limit to top 20 enriched terms by adjusted p-value
  top_terms <- gp_res %>%
    arrange(p_value) 
   # dplyr::slice_head(n = 20)

  # Define colour scale based on number of unique cpg_groupings
  n_sources <- length(unique(top_terms$source))
  fill_cols <- okabe_ito_palette(max(3, n_sources))  # ensure enough distinct colours

  # Generate enrichment plot
  organ_list[[organ]]$gp_plot <- ggplot(top_terms, aes(
      x = reorder(term_name, -log10(p_value)),
      y = -log10(p_value),
      fill = source
    )) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal(base_size = 12) +
    scale_fill_manual(values = fill_cols, name = "Database") +
    labs(
      title = paste("g:Profiler enrichment -", organ),
      x = NULL,
      y = expression(-log[10](adjusted~p~value))
    ) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  print(organ_list[[organ]]$gp_plot)
  
  ggsave(
  filename = paste0(output_directory, "/Functional_enrichment_universal_Multi_organ_GOterm_",organ,".png"),
  plot = organ_list[[organ]]$gp_plot,
  height = 6,
  width = 10,
  dpi = 400
)
    enriched_terms <- as.data.frame(organ_list[[organ]][["gp"]][["result"]])
enriched_terms$term_name

write_csv(enriched_terms,paste0("results/Profile_enrichment_Multi_organ_",organ,".csv"))
}

  }, error = function(e) {
    message("Processing failed for ", organ, ": ", e$message)
  })
}



```




## Universal all single organ overlapping cps 
```{r s, echo=T, message=FALSE, warning=FALSE, results='hide'}
###############################################################
# Epigenetic Clock CpG Functional Annotation and Enrichment
# Purpose: Identify functional enrichment (GO / KEGG) of CpGs 
#          contributing to best-performing organ-specific 
#          epigenetic clock models in Danio rerio.
###############################################################

#--------------------------------------------------------------
# 0. Load Required Packages
#--------------------------------------------------------------
suppressPackageStartupMessages({
library(missMethyl)             # CpG-centric enrichment (gometh)
library(dplyr)                  # Data manipulation
library(GenomicRanges)          # Genomic interval operations
library(TxDb.Drerio.UCSC.danRer11.refGene) # Zebrafish gene annotation
library(clusterProfiler)        # Classical GO/KEGG enrichment
library(org.Dr.eg.db)           # Zebrafish gene ontology database
library(gprofiler2)             # Orthogonal enrichment tool (g:Profiler)
  library(patchwork)
})
#--------------------------------------------------------------
# 1. Condense Model Metrics: Select the Best Model per Organ
#--------------------------------------------------------------
# 'merged_data$metrics' contains model performance metrics 
# (e.g., mean absolute error [MAE] and R² from training).



#--------------------------------------------------------------
# 2. Rank Models per Organ by Combined Performance
#--------------------------------------------------------------
# A composite ranking is computed as:
# - lower MAE = better
# - higher R² = better
# - combined rank = mae_rank + r2_rank
# The top-performing model per organ is retained.

#--------------------------------------------------------------
# 3. Extract CpGs from the Best-Performing Models
#--------------------------------------------------------------
# 'cpg_data_validation' contains model-specific CpGs; extract those
# corresponding to the top-ranked models per organ.

Multi_organ_best_cpg  <- unique(best_cpg_overlap_single)

#--------------------------------------------------------------
# 5. Extract Candidate and Selected CpGs
#--------------------------------------------------------------
# Candidate CpGs = all CpGs available for model training.
# Selected CpGs = CpGs retained in the final best model.
# Both are represented as genomic coordinate strings 
# (e.g., "chr1:24557370:24557869").

Multi_organ_candidate_cpg <- methylation_data_list$Multi_organ$Probe
Brain_candidate_cpg     <- methylation_data_list$Brain$Probe
Caudal_Fin_candidate_cpg       <- methylation_data_list$Caudal_Fin$Probe
Intestines_candidate_cpg <- methylation_data_list$Intestines$Probe

Multi_organ_selected_cpg  <- Multi_organ_best_cpg 
Brain_selected_cpg      <-Multi_organ_best_cpg 
Caudal_Fin_selected_cpg    <- Multi_organ_best_cpg 
Intestines_selected_cpg  <-Multi_organ_best_cpg 


# need to make sure cpgs match annotation_database 

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}

Multi_organ_candidate_cpg <-adding_chr_cpg(Multi_organ_candidate_cpg)
Brain_candidate_cpg     <-adding_chr_cpg(Brain_candidate_cpg  )
Caudal_Fin_candidate_cpg       <-adding_chr_cpg(Caudal_Fin_candidate_cpg   )
Intestines_candidate_cpg <-adding_chr_cpg(Intestines_candidate_cpg)

Multi_organ_selected_cpg  <- adding_chr_cpg(Multi_organ_selected_cpg)
Brain_selected_cpg      <-adding_chr_cpg(Brain_selected_cpg )
Caudal_Fin_selected_cpg <- adding_chr_cpg(Caudal_Fin_selected_cpg)
Intestines_selected_cpg  <-adding_chr_cpg(Intestines_selected_cpg)

#--------------------------------------------------------------
# 6. Functional Annotation Loop across Organs
#--------------------------------------------------------------
# For each organ, CpGs are:
#  (i) parsed into GRanges,
#  (ii) mapped to the nearest genes using TxDb,
#  (iii) subjected to multiple enrichment analyses:
#       - missMethyl::gometh (bias-corrected CpG enrichment)
#       - clusterProfiler::enrichGO (classical GO)
#       - gprofiler2::gost (independent validation)
# Results are stored in a nested list (organ_list).

organ_list <- list()
organs <- c( "Multi_organ","Brain", "Caudal_Fin" , "Intestines")

for (organ in organs) {
  message("\n--- Processing ", organ, " ---")

  organ_list[[organ]] <- list()

  tryCatch({

    #----------------------------------------------------------
    # 6.1 Retrieve CpG Vectors
    #----------------------------------------------------------
    if (!exists(paste0(organ, "_selected_cpg"))) stop("Missing selected CpG object")
    if (!exists(paste0(organ, "_candidate_cpg"))) stop("Missing candidate CpG object")

    selected_cpg <- get(paste0(organ, "_selected_cpg"))
    candidate_cpg <- get(paste0(organ, "_candidate_cpg"))

    message("Selected CpGs: ", length(selected_cpg),
            " | Candidate CpGs: ", length(candidate_cpg))

    #----------------------------------------------------------
    # 6.2 Parse Genomic Coordinates
    #----------------------------------------------------------
    # Each CpG name is split into chromosome, start, and end.
    selected_cpg_df <- data.frame(selected_cpgs = selected_cpg, stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpgs, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

    candidate_cpg_df <- data.frame(selected_cpgs = candidate_cpg, stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpgs, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

    #----------------------------------------------------------
    # 6.3 Construct GRanges Objects
    #----------------------------------------------------------
    selected_gr <- GRanges(seqnames = selected_cpg_df$chr,
                           ranges = IRanges(start = selected_cpg_df$start,
                                            end = selected_cpg_df$end))

    candidate_gr <- GRanges(seqnames = candidate_cpg_df$chr,
                            ranges = IRanges(start = candidate_cpg_df$start,
                                             end = candidate_cpg_df$end))

    #----------------------------------------------------------
    # 6.4 Map CpGs to the Nearest Genes
    #----------------------------------------------------------
    txdb <- TxDb.Drerio.UCSC.danRer11.refGene
    genes_gr <- genes(txdb)

    selected_entrez  <- names(genes_gr)[nearest(selected_gr, genes_gr)]
    candidate_entrez <- names(genes_gr)[nearest(candidate_gr, genes_gr)]

    # Store gene sets for manual inspection
    organ_list[[organ]]$selected_genes  <- selected_entrez
    organ_list[[organ]]$candidate_genes <- candidate_entrez

    message("Mapped ", length(selected_entrez), " selected CpGs to genes.")
    message("Mapped ", length(candidate_entrez), " candidate CpGs to genes.")

    #----------------------------------------------------------
    # 6.5 CpG-Level Enrichment using missMethyl (GO + KEGG)
    #----------------------------------------------------------
    organ_list[[organ]]$gst <- tryCatch({
      gometh(sig.cpg = selected_entrez,
             all.cpg = candidate_entrez,
             collection = "GO",
             plot.bias = TRUE)
    }, error = function(e) {
      message("gometh GO failed: ", e$message)
      NULL
    })

    organ_list[[organ]]$gst_kegg <- tryCatch({
      gometh(sig.cpg = selected_entrez,
             all.cpg = candidate_entrez,
             collection = "KEGG",
             plot.bias = TRUE)
    }, error = function(e) {
      message("gometh KEGG failed: ", e$message)
      NULL
    })

    #----------------------------------------------------------
    # 6.6 Classical GO Enrichment via clusterProfiler
    #----------------------------------------------------------
    message("Starting - GO enrichment")

    selected_genes  <- unique(selected_entrez)
    candidate_genes <- unique(candidate_entrez)

    organ_list[[organ]]$ego <- enrichGO(
      gene          = selected_genes,
      universe      = candidate_genes,
      OrgDb         = org.Dr.eg.db,
      keyType       = "ENTREZID",
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.2,
      readable      = TRUE
    )

    # Generate a dotplot if results exist
    if (nrow(as.data.frame(organ_list[[organ]]$ego)) > 0) {
      organ_list[[organ]]$enrich_dotplot <- dotplot(
        organ_list[[organ]]$ego,
        showCategory = 20
      )
    } else {
      message("No GO enrichment terms found for ", organ)
    }

    #----------------------------------------------------------
    # 6.7 Functional Enrichment with g:Profiler
    #----------------------------------------------------------
    warning("Starting - g:Profiler enrichment")

    organ_list[[organ]]$gp <- gost(
      query              = selected_genes,
      organism           = "drerio",
      correction_method  = "fdr",
      sources            = c("GO:BP", "GO:MF", "GO:CC", "KEGG")
    )

    # Visualise g:Profiler results if available
if (!is.null(organ_list[[organ]]$gp) && !is.null(organ_list[[organ]]$gp$result)) {

  # Store the result table separately for convenience
  gp_res <- organ_list[[organ]]$gp$result

  # Limit to top 20 enriched terms by adjusted p-value
  top_terms <- gp_res %>%
    arrange(p_value) 
   # dplyr::slice_head(n = 20)

  # Define colour scale based on number of unique cpg_groupings
  n_sources <- length(unique(top_terms$source))
  fill_cols <- okabe_ito_palette(max(3, n_sources))  # ensure enough distinct colours

  # Generate enrichment plot
  organ_list[[organ]]$gp_plot <- ggplot(top_terms, aes(
      x = reorder(term_name, -log10(p_value)),
      y = -log10(p_value),
      fill = source
    )) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal(base_size = 12) +
    scale_fill_manual(values = fill_cols, name = "Database") +
    labs(
      title = paste("g:Profiler enrichment - Soma"),
      x = NULL,
      y = expression(-log[10](adjusted~p~value))
    ) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )

  print(organ_list[[organ]]$gp_plot)
  
  ggsave(
  filename = paste0(output_directory, "/Functional_enrichment_universal_soma_overlap_GOterm_",organ,".png"),
  plot = organ_list[[organ]]$gp_plot,
  height = 6,
  width = 10,
  dpi = 400
)
      enriched_terms <- as.data.frame(organ_list[[organ]][["gp"]][["result"]])
enriched_terms$term_name

write_csv(enriched_terms,paste0("results/Profile_enrichment_soma_overlap_",organ,".csv"))
}

  }, error = function(e) {
    message("Processing failed for ", organ, ": ", e$message)
  })
}

functional_profiling_overlap_soma <- organ_list$Multi_organ$gp_plot

```




#~


## Genomic context of selected CpGs

### Organ specific 
Genomic context ( promoter, gene body, enhancer region) and CpG island features - https://doi.org/10.1038/s43856-025-00739-4

Merge CpGs used in clocks with gene and CpG island annotation:


```{r t}

#============================================================
# Epigenetic Clock CpG Annotation Workflow – Zebrafish
#============================================================
# Purpose: Identify best-performing CpGs from organ-specific 
# epigenetic clock models and annotate them with genomic context
# (promoter, gene body, nearest gene).
#------------------------------------------------------------
# Dependencies:
#   - GenomicRanges
#   - dplyr
#   - stringr
#   - tidyr
#   - GenomicFeatures
#   - TxDb.Drerio.UCSC.danRer11.refGene
#============================================================

# Load required packages
suppressPackageStartupMessages({
library(GenomicRanges)        # GRanges object for genomic ranges
library(dplyr)                # Data manipulation
library(stringr)              # String parsing
library(tidyr)                # Data tidying (separate)
library(GenomicFeatures)      # Extract genomic features
library(TxDb.Drerio.UCSC.danRer11.refGene)  # Zebrafish gene annotation
  library(patchwork)
})
#============================================================
# 1. Condense epigenetic clock metrics to select best model per organ
#------------------------------------------------------------
# metrics_data_validationcontains model performance metrics (MAE, R²)

#============================================================
# 3. Extract CpGs from best-performing models
#------------------------------------------------------------
Multi_organ_best_cpg <- tibble::tibble(
  organ = names(best_selected_cpgs_organs),
  cpgs  = best_selected_cpgs_organs
) |>
  tidyr::unnest_longer(cpgs, values_to = "cpg")

#============================================================
# 4. Parse CpG genomic coordinates
#------------------------------------------------------------
# CpGs are stored as "chr:start:end" strings
Multi_organ_best_cpg_parsed <- data.frame( selected_cpg = Multi_organ_best_cpg$cpg  , stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpg, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end),
             organ = Multi_organ_best_cpg$organ)

#============================================================
# 5. Create GRanges object for CpGs
#------------------------------------------------------------
cpg_gr <- GRanges(
  seqnames = Multi_organ_best_cpg_parsed$chr,        # chromosome
  ranges = IRanges(start = Multi_organ_best_cpg_parsed$start,
                   end   = Multi_organ_best_cpg_parsed$end),  # coordinates

)

#============================================================
# 6. Load zebrafish gene annotation (TxDb)
#------------------------------------------------------------
txdb <- TxDb.Drerio.UCSC.danRer11.refGene

#============================================================
# 7. Extract genomic features for annotation
#------------------------------------------------------------
# Promoter regions: 2 kb upstream, 500 bp downstream of TSS
promoters_gr <- promoters(genes(txdb), upstream = 2000, downstream = 500)

# Gene bodies
genes_gr <- genes(txdb)

#============================================================
# 8. Annotate CpGs with promoter and gene body overlap
#------------------------------------------------------------
# Overlaps with promoters
hits_promoters <- findOverlaps(cpg_gr, promoters_gr)
cpg_gr$in_promoter <- FALSE
cpg_gr$in_promoter[queryHits(hits_promoters)] <- TRUE

# Overlaps with gene bodies
hits_genes <- findOverlaps(cpg_gr, genes_gr)
cpg_gr$in_gene <- FALSE
cpg_gr$in_gene[queryHits(hits_genes)] <- TRUE

#============================================================
# 9. Annotate nearest gene for each CpG
#------------------------------------------------------------
nearest_genes <- nearest(cpg_gr, genes_gr)
cpg_gr$nearest_gene_id <- names(genes_gr)[nearest_genes]

#============================================================
# 10. Summarise CpG distribution per organ
#------------------------------------------------------------
summary_df <- as.data.frame(cpg_gr) %>%
  mutate(organ = Multi_organ_best_cpg$organ) %>%
  group_by(organ)%>%
  summarise(
    total_cpgs   = n(),                            # total CpGs in best model
    in_promoter   = sum(in_promoter),              # CpGs overlapping promoters
    in_gene       = sum(in_gene),                  # CpGs overlapping gene bodies
    unique_genes  = n_distinct(nearest_gene_id)    # distinct nearest genes
  )

# Print summary
print(summary_df)

```
This table shows the genomic distribution of CpGs selected by the best-performing epigenetic clock models for each organ and across all organs:
total_cpgs: Number of CpGs in each organ-specific or universal model.
in_promoter: CpGs overlapping promoter regions (−2 kb to +500 bp around TSS), which are likely regulatory.
in_gene: CpGs located within gene bodies, potentially affecting transcriptional regulation.
unique_genes: Number of distinct genes nearest to the selected CpGs.

Key points:
Brain has the largest model (48 CpGs) with many in promoters and gene bodies, reflecting complex tissue-specific methylation.
Caudal_Fin and Intestines have fewer CpGs, suggesting sparse but efficient predictors.
The universal “Multi_organ” model largely avoids promoters, with CpGs primarily in gene bodies or intergenic regions, highlighting shared ageing signals across tissues.

#### Plot for genomic context 

```{r u}
#============================================================
# Plot : Genomic Distribution of CpGs
#============================================================
# Purpose:
#   Visualise the distribution of CpGs (total, promoter, gene body)
#   for best organ-specific epigenetic clock models.
#   Colours are set using the Okabe-Ito colourblind-friendly palette.
#------------------------------------------------------------
# Dependencies:
#   - ggplot2
#   - tidyr
#   - dplyr
#============================================================
suppressPackageStartupMessages({
library(ggplot2)  # Core plotting library
library(tidyr)    # For reshaping data
library(dplyr)    # Data manipulation
})
#------------------------------------------------------------
# 1. Prepare data in long format for ggplot2
#------------------------------------------------------------
# summary_df columns:
#   - organ: Organ name
#   - total_cpgs: Number of CpGs in best model
#   - in_promoter: Number of CpGs overlapping promoters
#   - in_gene: Number of CpGs overlapping gene bodies
plot_df <- summary_df %>%
 # filter(organ != "Multi_organ") %>%
  dplyr::select(organ, total_cpgs, in_promoter, in_gene) %>%
  pivot_longer(
    cols = c(total_cpgs, in_promoter, in_gene),  # variables to pivot
    names_to = "feature",                         # new column identifying feature type
    values_to = "count"                           # new column with counts
  ) %>%
  mutate(
    feature = factor(feature, levels = c(  "Total CpGs" = "total_cpgs",
                                           "Promoter CpGs" = "in_promoter",
                                          "Genebody CpGs" = "in_gene" 
 ))
  )


#------------------------------------------------------------
# 3. Create bar plot
#------------------------------------------------------------
p <- ggplot(plot_df, aes(x = organ, y = count, fill = feature)) +
  geom_bar(stat = "identity", position = "dodge") +
  
  scale_fill_manual(
    values = c(
      total_cpgs  = okabe_ito[1],
      in_promoter = okabe_ito[2],
      in_gene     = okabe_ito[3]
    ),
    labels = c(
      total_cpgs  = "Total CpGs",
      in_promoter = "Promoter CpGs",
      in_gene     = "Genebody CpGs"
    )
  ) +

  labs(
    title = "Genomic Distribution of selected CpGs in Best Clock Models",
    x = "",
    y = "Number of CpGs",
    fill = "Feature"
  ) +
  
  plot_theme 

# Display plot
p

#------------------------------------------------------------
# 4. Save figure to file
#------------------------------------------------------------
# Parameters: height = 6in, width = 6in, dpi = 400 (high-resolution for publication)
ggsave(
  filename = paste0(output_directory, "/Genomic_Distribution_organ_specific.png"),
  plot = p,
  height = 6,
  width = 8,
  dpi = 400
)

```



### Universal all models 
Genomic context ( promoter, gene body, enhancer region) and CpG island features - https://doi.org/10.1038/s43856-025-00739-4

Merge CpGs used in clocks with gene and CpG island annotation:


```{r v}

#============================================================
# Epigenetic Clock CpG Annotation Workflow – Zebrafish
#============================================================
# Purpose: Identify best-performing CpGs from organ-specific 
# epigenetic clock models and annotate them with genomic context
# (promoter, gene body, nearest gene).
#------------------------------------------------------------
# Dependencies:
#   - GenomicRanges
#   - dplyr
#   - stringr
#   - tidyr
#   - GenomicFeatures
#   - TxDb.Drerio.UCSC.danRer11.refGene
#============================================================

# Load required packages
suppressPackageStartupMessages({
library(GenomicRanges)        # GRanges object for genomic ranges
library(dplyr)                # Data manipulation
library(stringr)              # String parsing
library(tidyr)                # Data tidying (separate)
library(GenomicFeatures)      # Extract genomic features
library(TxDb.Drerio.UCSC.danRer11.refGene)  # Zebrafish gene annotation
})
#============================================================
# 1. Condense epigenetic clock metrics to select best model per organ
#------------------------------------------------------------
# metrics_data_validationcontains model performance metrics (MAE, R²)

#============================================================
# 3. Extract CpGs from best-performing models
#------------------------------------------------------------

Multi_organ_best_cpg  <- unique(best_cpg_overlap_all)
adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}


Multi_organ_best_cpg <- adding_chr_cpg(Multi_organ_best_cpg)

#============================================================
# 4. Parse CpG genomic coordinates
#------------------------------------------------------------
# CpGs are stored as "chr:start:end" strings
Multi_organ_best_cpg_parsed <- data.frame( selected_cpg = Multi_organ_best_cpg , stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpg, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

#============================================================
# 5. Create GRanges object for CpGs
#------------------------------------------------------------
cpg_gr <- GRanges(
  seqnames = Multi_organ_best_cpg_parsed$chr,        # chromosome
  ranges = IRanges(start = Multi_organ_best_cpg_parsed$start,
                   end   = Multi_organ_best_cpg_parsed$end),  # coordinates

)

#============================================================
# 6. Load zebrafish gene annotation (TxDb)
#------------------------------------------------------------
txdb <- TxDb.Drerio.UCSC.danRer11.refGene

#============================================================
# 7. Extract genomic features for annotation
#------------------------------------------------------------
# Promoter regions: 2 kb upstream, 500 bp downstream of TSS
promoters_gr <- promoters(genes(txdb), upstream = 2000, downstream = 500)

# Gene bodies
genes_gr <- genes(txdb)

#============================================================
# 8. Annotate CpGs with promoter and gene body overlap
#------------------------------------------------------------
# Overlaps with promoters
hits_promoters <- findOverlaps(cpg_gr, promoters_gr)
cpg_gr$in_promoter <- FALSE
cpg_gr$in_promoter[queryHits(hits_promoters)] <- TRUE

# Overlaps with gene bodies
hits_genes <- findOverlaps(cpg_gr, genes_gr)
cpg_gr$in_gene <- FALSE
cpg_gr$in_gene[queryHits(hits_genes)] <- TRUE

#============================================================
# 9. Annotate nearest gene for each CpG
#------------------------------------------------------------
nearest_genes <- nearest(cpg_gr, genes_gr)
cpg_gr$nearest_gene_id <- names(genes_gr)[nearest_genes]

#============================================================
# 10. Summarise CpG distribution per organ
#------------------------------------------------------------
summary_df <- as.data.frame(cpg_gr) %>%
  summarise(
    total_cpgs   = n(),                            # total CpGs in best model
    in_promoter   = sum(in_promoter),              # CpGs overlapping promoters
    in_gene       = sum(in_gene),                  # CpGs overlapping gene bodies
    unique_genes  = n_distinct(nearest_gene_id)    # distinct nearest genes
  )

# Print summary
print(summary_df)

```
This table shows the genomic distribution of CpGs selected by the best-performing epigenetic clock models for each organ and across all organs:
total_cpgs: Number of CpGs in each organ-specific or universal model.
in_promoter: CpGs overlapping promoter regions (−2 kb to +500 bp around TSS), which are likely regulatory.
in_gene: CpGs located within gene bodies, potentially affecting transcriptional regulation.
unique_genes: Number of distinct genes nearest to the selected CpGs.

Key points:
Brain has the largest model (48 CpGs) with many in promoters and gene bodies, reflecting complex tissue-specific methylation.
Caudal_Fin and Intestines have fewer CpGs, suggesting sparse but efficient predictors.
The universal “Multi_organ” model largely avoids promoters, with CpGs primarily in gene bodies or intergenic regions, highlighting shared ageing signals across tissues.

#### Plot for genomic context 

```{r w}
#============================================================
# Plot : Genomic Distribution of CpGs
#============================================================
# Purpose:
#   Visualise the distribution of CpGs (total, promoter, gene body)
#   for best organ-specific epigenetic clock models.
#   Colours are set using the Okabe-Ito colourblind-friendly palette.
#------------------------------------------------------------
# Dependencies:
#   - ggplot2
#   - tidyr
#   - dplyr
#============================================================
suppressPackageStartupMessages({
library(ggplot2)  # Core plotting library
library(tidyr)    # For reshaping data
library(dplyr)    # Data manipulation
})
#------------------------------------------------------------
# 1. Prepare data in long format for ggplot2
#------------------------------------------------------------
# summary_df columns:
#   - organ: Organ name
#   - total_cpgs: Number of CpGs in best model
#   - in_promoter: Number of CpGs overlapping promoters
#   - in_gene: Number of CpGs overlapping gene bodies
plot_df <- summary_df %>%
 # filter(organ != "Multi_organ") %>%
  dplyr::select(total_cpgs, in_promoter, in_gene) %>%
  pivot_longer(
    cols = c(total_cpgs, in_promoter, in_gene),  # variables to pivot
    names_to = "feature",                         # new column identifying feature type
    values_to = "count"                           # new column with counts
  ) %>%
  mutate(organ = "Soma")

#------------------------------------------------------------
# 2. Define colours using Okabe-Ito palette
#------------------------------------------------------------
# Assign first three colours to the three features
feature_colors <- okabe_ito[1:3]

#------------------------------------------------------------
# 3. Create bar plot
#------------------------------------------------------------
p <- ggplot(plot_df, aes(x = organ, y = count, fill = feature)) +
  geom_bar(stat = "identity", position = "dodge") +
  
  scale_fill_manual(
    values = c(
      total_cpgs  = okabe_ito[1],
      in_promoter = okabe_ito[2],
      in_gene     = okabe_ito[3]
    ),
    labels = c(
      total_cpgs  = "Total CpGs",
      in_promoter = "Promoter CpGs",
      in_gene     = "Genebody CpGs"
    )
  ) +

  labs(
    title = "Genomic Distribution of selected CpGs in Best Clock Models",
    x = "",
    y = "Number of CpGs",
    fill = "Feature"
  ) +
  theme_classic2(base_size = 16)   +
  plot_theme 

# Display plot
p


#------------------------------------------------------------
# 4. Save figure to file
#------------------------------------------------------------
# Parameters: height = 6in, width = 6in, dpi = 400 (high-resolution for publication)
ggsave(
  filename = paste0(output_directory, "/Genomic_Distribution_universal_all_models.png"),
  plot = p,
  height = 6,
  width = 8,
  dpi = 400
)

```



### Universal overlapping soma  
Genomic context ( promoter, gene body, enhancer region) and CpG island features - https://doi.org/10.1038/s43856-025-00739-4

Merge CpGs used in clocks with gene and CpG island annotation:


```{r x}

#============================================================
# Epigenetic Clock CpG Annotation Workflow – Zebrafish
#============================================================
# Purpose: Identify best-performing CpGs from organ-specific 
# epigenetic clock models and annotate them with genomic context
# (promoter, gene body, nearest gene).
#------------------------------------------------------------
# Dependencies:
#   - GenomicRanges
#   - dplyr
#   - stringr
#   - tidyr
#   - GenomicFeatures
#   - TxDb.Drerio.UCSC.danRer11.refGene
#============================================================

# Load required packages
suppressPackageStartupMessages({
library(GenomicRanges)        # GRanges object for genomic ranges
library(dplyr)                # Data manipulation
library(stringr)              # String parsing
library(tidyr)                # Data tidying (separate)
library(GenomicFeatures)      # Extract genomic features
library(TxDb.Drerio.UCSC.danRer11.refGene)  # Zebrafish gene annotation
})
#============================================================
# 1. Condense epigenetic clock metrics to select best model per organ
#------------------------------------------------------------
# metrics_data_validationcontains model performance metrics (MAE, R²)

#============================================================
# 3. Extract CpGs from best-performing models
#------------------------------------------------------------

Multi_organ_best_cpg  <- unique(best_cpg_overlap_single)

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}


Multi_organ_best_cpg <- adding_chr_cpg(Multi_organ_best_cpg)



#============================================================
# 4. Parse CpG genomic coordinates
#------------------------------------------------------------
# CpGs are stored as "chr:start:end" strings
Multi_organ_best_cpg_parsed <- data.frame( selected_cpg = Multi_organ_best_cpg , stringsAsFactors = FALSE) %>%
      tidyr::separate(selected_cpg, into = c("chr", "start", "end"),
                      sep = ":", remove = FALSE, fill = "right") %>%
      mutate(start = as.numeric(start), end = as.numeric(end))

#============================================================
# 5. Create GRanges object for CpGs
#------------------------------------------------------------
cpg_gr <- GRanges(
  seqnames = Multi_organ_best_cpg_parsed$chr,        # chromosome
  ranges = IRanges(start = Multi_organ_best_cpg_parsed$start,
                   end   = Multi_organ_best_cpg_parsed$end),  # coordinates

)

#============================================================
# 6. Load zebrafish gene annotation (TxDb)
#------------------------------------------------------------
txdb <- TxDb.Drerio.UCSC.danRer11.refGene

#============================================================
# 7. Extract genomic features for annotation
#------------------------------------------------------------
# Promoter regions: 2 kb upstream, 500 bp downstream of TSS
promoters_gr <- promoters(genes(txdb), upstream = 2000, downstream = 500)

# Gene bodies
genes_gr <- genes(txdb)

#============================================================
# 8. Annotate CpGs with promoter and gene body overlap
#------------------------------------------------------------
# Overlaps with promoters
hits_promoters <- findOverlaps(cpg_gr, promoters_gr)
cpg_gr$in_promoter <- FALSE
cpg_gr$in_promoter[queryHits(hits_promoters)] <- TRUE

# Overlaps with gene bodies
hits_genes <- findOverlaps(cpg_gr, genes_gr)
cpg_gr$in_gene <- FALSE
cpg_gr$in_gene[queryHits(hits_genes)] <- TRUE

#============================================================
# 9. Annotate nearest gene for each CpG
#------------------------------------------------------------
nearest_genes <- nearest(cpg_gr, genes_gr)
cpg_gr$nearest_gene_id <- names(genes_gr)[nearest_genes]

#============================================================
# 10. Summarise CpG distribution per organ
#------------------------------------------------------------
summary_df <- as.data.frame(cpg_gr) %>%
  summarise(
    total_cpgs   = n(),                            # total CpGs in best model
    in_promoter   = sum(in_promoter),              # CpGs overlapping promoters
    in_gene       = sum(in_gene),                  # CpGs overlapping gene bodies
    unique_genes  = n_distinct(nearest_gene_id)    # distinct nearest genes
  )

# Print summary
print(summary_df)

```
This table shows the genomic distribution of CpGs selected by the best-performing epigenetic clock models for each organ and across all organs:
total_cpgs: Number of CpGs in each organ-specific or universal model.
in_promoter: CpGs overlapping promoter regions (−2 kb to +500 bp around TSS), which are likely regulatory.
in_gene: CpGs located within gene bodies, potentially affecting transcriptional regulation.
unique_genes: Number of distinct genes nearest to the selected CpGs.

Key points:
Brain has the largest model (48 CpGs) with many in promoters and gene bodies, reflecting complex tissue-specific methylation.
Caudal_Fin and Intestines have fewer CpGs, suggesting sparse but efficient predictors.
The universal “Multi_organ” model largely avoids promoters, with CpGs primarily in gene bodies or intergenic regions, highlighting shared ageing signals across tissues.

#### Plot for genomic context 

```{r y}
#============================================================
# Plot : Genomic Distribution of CpGs
#============================================================
# Purpose:
#   Visualise the distribution of CpGs (total, promoter, gene body)
#   for best organ-specific epigenetic clock models.
#   Colours are set using the Okabe-Ito colourblind-friendly palette.
#------------------------------------------------------------
# Dependencies:
#   - ggplot2
#   - tidyr
#   - dplyr
#============================================================
suppressPackageStartupMessages({
library(ggplot2)  # Core plotting library
library(tidyr)    # For reshaping data
library(dplyr)    # Data manipulation
})
#------------------------------------------------------------
# 1. Prepare data in long format for ggplot2
#------------------------------------------------------------
# summary_df columns:
#   - organ: Organ name
#   - total_cpgs: Number of CpGs in best model
#   - in_promoter: Number of CpGs overlapping promoters
#   - in_gene: Number of CpGs overlapping gene bodies
plot_df <- summary_df %>%
 # filter(organ != "Multi_organ") %>%
  dplyr::select(total_cpgs, in_promoter, in_gene) %>%
  pivot_longer(
    cols = c(total_cpgs, in_promoter, in_gene),  # variables to pivot
    names_to = "feature",                         # new column identifying feature type
    values_to = "count"                           # new column with counts
  ) %>%
  mutate(organ = "Soma")

#------------------------------------------------------------
# 2. Define colours using Okabe-Ito palette
#------------------------------------------------------------
# Assign first three colours to the three features
feature_colors <- okabe_ito[1:3]

#------------------------------------------------------------
# 3. Create bar plot
#------------------------------------------------------------
p <- ggplot(plot_df, aes(x = organ, y = count, fill = feature)) +
  geom_bar(stat = "identity", position = "dodge") +
  
  scale_fill_manual(
    values = c(
      total_cpgs  = okabe_ito[1],
      in_promoter = okabe_ito[2],
      in_gene     = okabe_ito[3]
    ),
    labels = c(
      total_cpgs  = "Total CpGs",
      in_promoter = "Promoter CpGs",
      in_gene     = "Genebody CpGs"
    )
  ) +

  labs(
    title = "Genomic Distribution of selected CpGs",
    x = "",
    y = "Number of CpGs",
    fill = "Feature"
  ) +
  theme_classic2(base_size = 16)   +
  plot_theme 

# Display plot
p
genomic_context_overlap_soma <-p 
#------------------------------------------------------------
# 4. Save figure to file
#------------------------------------------------------------
# Parameters: height = 6in, width = 6in, dpi = 400 (high-resolution for publication)
ggsave(
  filename = paste0(output_directory, "/Genomic_Distribution_universal_soma_overlap.png"),
  plot = genomic_context_overlap_soma,
  height = 6,
  width = 8,
  dpi = 400
)

```


# ~ 

