---
title: "Cross-Tissue analysis - part 7: methylation_drift_and_pcgt_enrichment"
author: "Jean-Charles de Coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
PIPELINE STAGE: 06 / Aim 2: Cross-Tissue Ageing Biology - part 7 of 10
=================================================================
Evolutionary/developmental framing; directionality of age-related methylation drift (hyper- vs hypomethylation) per tissue; Polycomb Group Target (PCGT) gene enrichment for three overlapping-CpG definitions.

Depends on objects created by part 6 (06_cpg_functional_annotation_enrichment.Rmd) having already been run in the same R
session (or knitted first, with its environment preserved) - this file does not reload raw data from scratch.
=================================================================
-->

```{r config_part7}
# Set the root directory of this project (edit here, or export
# EPICLOCK_ROOT before running).
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
```

# Evolutionary and developmental perspectives 

CpG overlap with  proliferative methylation patterns ( eg PCGTs unmethylated in development byt hypermethylated with age) - developmental ageing trajectories accross organs - https://doi.org/10.1038/s43856-025-00739-4

Directionality of methylation drift - does it align with model of systemic ageing vs organ turnover differences - https://doi.org/10.7554/eLife.40675

 Identify enrichment for Polycomb target genes and developmental loci

This checks whether your clock CpGs are enriched near Polycomb group target genes (PCGTs) or developmental regulators.

 using PCGT list from humans from Bracken et al. (2006) - 10.1101/gad.381706
 ### Pre-requisit - need to extract gene informaiton from table 
 
### Organ specific - repeatidy selecting cpgs 
```{r z}
##############################################################
# Zebrafish CpG Model Analysis & PCGT Enrichment Pipeline
# Description: Select best CpG-based epigenetic clock models 
# per organ, map CpGs to genes, convert human PCGT genes to 
# zebrafish orthologues, and test enrichment.
##############################################################

#--------------------------------------------------------------
# 0. Load Required Libraries
#--------------------------------------------------------------
suppressPackageStartupMessages({
library(dplyr)                  
library(tidyr)                  
library(GenomicRanges)          
library(TxDb.Drerio.UCSC.danRer11.refGene) 
library(clusterProfiler)        
library(org.Dr.eg.db)           
library(missMethyl)             
library(gprofiler2)             
library(biomaRt)                
library(ggplot2)
  library(readr)
})
#--------------------------------------------------------------
# 1. Load Data
#--------------------------------------------------------------
# merged_data$metrics: performance metrics (MAE, R²) per model
# cpg_data_validation: CpGs per model
# best_model_*: fitted model objects per organ
# Bracken2006_PCGT_human: human PCGT gene list


Bracken2006_PCGT_human <- read_csv(here("data/Bracken_et_al_2006_PCGT_list/Bracken2006_PCCT_human.csv"))

#--------------------------------------------------------------
# 3. Extract CpGs from Best Models
#--------------------------------------------------------------
Multi_organ_best_cpg <- tibble::tibble(
  organ = names(best_selected_cpgs_organs),
  cpgs  = best_selected_cpgs_organs
) |>
  tidyr::unnest_longer(cpgs, values_to = "cpg")

Multi_organ_best_cpg  <- Multi_organ_best_cpg %>%
  pull(unique(cpg))


meta_data <- list(
  Multi_organ      = best_model_list$Multi_organ$sample_data_filtered,
  Brain           = best_model_list$Brain$sample_data_filtered,
  Caudal_Fin = best_model_list$Caudal_Fin$sample_data_filtered,
  Intestines      = best_model_list$Intestines$sample_data_filtered
)


#--------------------------------------------------------------
# 5. Extract Candidate and Selected CpGs
#--------------------------------------------------------------
# Candidate CpGs = all CpGs available for model training.
# Selected CpGs = CpGs retained in the final best model.
# Both are represented as genomic coordinate strings 
# (e.g., "chr1:24557370:24557869").

Multi_organ_candidate_cpg <- methylation_data_list$Multi_organ$Probe
Brain_candidate_cpg     <- methylation_data_list$Brain$Probe
Caudal_Fin_candidate_cpg       <- methylation_data_list$Caudal_Fin$Probe
Intestines_candidate_cpg <- methylation_data_list$Intestines$Probe

Multi_organ_selected_cpg  <- best_selected_cpgs_organs$Multi_organ
Brain_selected_cpg      <-best_selected_cpgs_organs$Brain
Caudal_Fin_selected_cpg        <-best_selected_cpgs_organs$Caudal_Fin
Intestines_selected_cpg  <-best_selected_cpgs_organs$Intestines


# need to make sure cpgs match annotation_database 

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}

Multi_organ_candidate_cpg <-adding_chr_cpg(Multi_organ_candidate_cpg)
Brain_candidate_cpg     <-adding_chr_cpg(Brain_candidate_cpg  )
Caudal_Fin_candidate_cpg       <-adding_chr_cpg(Caudal_Fin_candidate_cpg   )
Intestines_candidate_cpg <-adding_chr_cpg(Intestines_candidate_cpg)

Multi_organ_selected_cpg  <- adding_chr_cpg(Multi_organ_selected_cpg)
Brain_selected_cpg      <-adding_chr_cpg(Brain_selected_cpg )
Caudal_Fin_selected_cpg <- adding_chr_cpg(Caudal_Fin_selected_cpg)
Intestines_selected_cpg  <-adding_chr_cpg(Intestines_selected_cpg)


#--------------------------------------------------------------
# 4. parse selected cpgs 
#--------------------------------------------------------------
organs <- c("Multi_organ", "Brain", "Caudal_Fin" , "Intestines")
organ_list <- list ()

for (organ in organs) {
 
  message("Starting organ :", organ)
  
      selected_cpg <- get(paste0(organ, "_selected_cpg"))
    candidate_cpg <- get(paste0(organ, "_candidate_cpg"))
  
  # Function to ensure "chr" prefix
ensure_chr_prefix <- function(cpg_vector) {
  # Only add 'chr' if not already present
  cpg_vector <- ifelse(grepl("^chr", cpg_vector), cpg_vector, paste0("chr", cpg_vector))
  return(cpg_vector)
}

selected_cpg  <- ensure_chr_prefix(selected_cpg)
candidate_cpg <- ensure_chr_prefix(candidate_cpg)

  #----------------------------------------------------------
  # Parse genomic coordinates
  #----------------------------------------------------------
  selected_df <- data.frame(cpg = selected_cpg, stringsAsFactors = FALSE) %>%
    tidyr::separate(cpg, into = c("chr","start","end"), sep=":", fill="right") %>%
    mutate(start = as.numeric(start), end = as.numeric(end))
  
  candidate_df <- data.frame(cpg = candidate_cpg, stringsAsFactors = FALSE) %>%
    tidyr::separate(cpg, into = c("chr","start","end"), sep=":", fill="right") %>%
    mutate(start = as.numeric(start), end = as.numeric(end))
  
  #----------------------------------------------------------
  # Map to nearest genes using GRanges
  #----------------------------------------------------------
  selected_gr <- GRanges(seqnames = selected_df$chr,
                         ranges = IRanges(start=selected_df$start, end=selected_df$end))
  candidate_gr <- GRanges(seqnames = candidate_df$chr,
                          ranges = IRanges(start=candidate_df$start, end=candidate_df$end))
  
  genes_gr <- genes(TxDb.Drerio.UCSC.danRer11.refGene)
  selected_df$entrez  <- names(genes_gr)[nearest(selected_gr, genes_gr)]
  candidate_df$entrez <- names(genes_gr)[nearest(candidate_gr, genes_gr)]
  
  # Store per organ
  organ_list[[organ]] <- list(
    selected_cpg_df = selected_df,
    candidate_cpg_df = candidate_df
  )
}

#--------------------------------------------------------------
# 5. Convert Human PCGT Genes to Zebrafish Entrez IDs
#--------------------------------------------------------------
human <- useEnsembl("ensembl", dataset="hsapiens_gene_ensembl", mirror="useast")
pcgt_genes_human <- unique(na.omit(Bracken2006_PCGT_human$Gene))

# Get zebrafish orthologue Ensembl IDs
ortholog_table <- getBM(
  attributes=c("drerio_homolog_ensembl_gene","drerio_homolog_associated_gene_name"),
  filters="hgnc_symbol",
  values=pcgt_genes_human,
  mart=human
)

zfish_ensembl <- unique(ortholog_table$drerio_homolog_ensembl_gene)

# Map zebrafish Ensembl IDs to Entrez IDs
zebrafish <- useEnsembl("ensembl", dataset="drerio_gene_ensembl", mirror="useast")
zfish_mapping <- getBM(
  attributes=c("ensembl_gene_id","entrezgene_id"),
  filters="ensembl_gene_id",
  values=zfish_ensembl,
  mart=zebrafish
)

pcgt_zfish_entrez <- unique(na.omit(zfish_mapping$entrezgene_id))

#--------------------------------------------------------------
# 6. Compute PCGT Enrichment per Organ
#--------------------------------------------------------------
enrichment_results <- lapply(names(organ_list), function(org) {
  message("Processing organ: ", org)
  
  tryCatch({
    

  selected_df  <- organ_list[[org]]$selected_cpg_df
  candidate_df <- organ_list[[org]]$candidate_cpg_df
  
  all_genes <- unique(c(candidate_df$entrez, selected_df$entrez))
  pcgt_flag <- all_genes %in% pcgt_zfish_entrez
  clock_flag <- all_genes %in% selected_df$entrez
  
  contingency <- table(pcgt_flag, clock_flag)
  fisher_res <- fisher.test(contingency)
  
  data.frame(
    organ = org,
    n_selected = nrow(selected_df),
    n_candidate = nrow(candidate_df),
    overlap_n = sum(selected_df$entrez %in% pcgt_zfish_entrez),
    overlap_prop = mean(selected_df$entrez %in% pcgt_zfish_entrez),
    p_value = fisher_res$p.value,
    odds_ratio = fisher_res$estimate
  )
  } , error = function(e) {
    message("Processing failed for ", organ, ": ", e$message)
    })
  
})

enrichment_df <- do.call(rbind, enrichment_results)

#--------------------------------------------------------------
# 7. Plot PCGT Enrichment
#--------------------------------------------------------------
# Create significance labels

enrichment_df <- enrichment_df %>%
  mutate(sig_label = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    p_value < 0.10  ~ ".",
    TRUE            ~ ""
  ))

# Compute standard error for error bars
enrichment_df <- enrichment_df %>%
  rowwise() %>%
  mutate(
    se = sqrt(overlap_prop * (1 - overlap_prop) / n_selected),
    ymin = overlap_prop - se,
    ymax = overlap_prop + se,
    sig_y = ymax + 0.05  # y-position for significance label above error bar
  ) 

# Plot with significance labels
p <- ggplot(enrichment_df, aes(x=organ, y=overlap_prop, color=organ)) +
  geom_point(size=4) +
  geom_errorbar(aes(ymin=ymin, ymax=ymax), width=0.2) +
  geom_label(aes(label=sprintf("%.2f", overlap_prop)),
             vjust=-0.5, size=4, fill="white", color="black") +
  geom_text(aes(y=sig_y, label=sig_label), vjust=0, size=5, color="black") +
  labs(title="PCGT Orthologue Enrichment per Organ",
       y="Proportion of CpGs near PCGT orthologues",
       x=NULL) +
  scale_color_manual(values=okabe_ito_palette(4)) +
  
  theme(legend.position="none")  +
  plot_theme 

print(p)

ggsave(
  filename = file.path(output_directory, "PCGT_genes_organ_specific.png"),
  plot = p,
  height = 7,
  width = 7,
  dpi = 400
)

```

### Universal all models 
```{r aa}
##############################################################
# Zebrafish CpG Model Analysis & PCGT Enrichment Pipeline
# Description: Select best CpG-based epigenetic clock models 
# per organ, map CpGs to genes, convert human PCGT genes to 
# zebrafish orthologues, and test enrichment.
##############################################################

#--------------------------------------------------------------
# 0. Load Required Libraries
#--------------------------------------------------------------
suppressPackageStartupMessages({
library(dplyr)                  
library(tidyr)                  
library(GenomicRanges)          
library(TxDb.Drerio.UCSC.danRer11.refGene) 
library(clusterProfiler)        
library(org.Dr.eg.db)           
library(missMethyl)             
library(gprofiler2)             
library(biomaRt)                
library(ggplot2)
  library(readr)
})
#--------------------------------------------------------------
# 1. Load Data
#--------------------------------------------------------------
# merged_data$metrics: performance metrics (MAE, R²) per model
# cpg_data_validation: CpGs per model
# best_model_*: fitted model objects per organ
# Bracken2006_PCGT_human: human PCGT gene list


Bracken2006_PCGT_human <- read_csv(here("data/Bracken_et_al_2006_PCGT_list/Bracken2006_PCCT_human.csv"))

#--------------------------------------------------------------
# 3. Extract CpGs from Best Models
#--------------------------------------------------------------

Multi_organ_best_cpg  <- unique(best_cpg_overlap_all)


meta_data <- list(
  Multi_organ      = best_model_list$Multi_organ$sample_data_filtered,
  Brain           = best_model_list$Brain$sample_data_filtered,
  Caudal_Fin = best_model_list$Caudal_Fin$sample_data_filtered,
  Intestines      = best_model_list$Intestines$sample_data_filtered
)


#--------------------------------------------------------------
# 5. Extract Candidate and Selected CpGs
#--------------------------------------------------------------
# Candidate CpGs = all CpGs available for model training.
# Selected CpGs = CpGs retained in the final best model.
# Both are represented as genomic coordinate strings 
# (e.g., "chr1:24557370:24557869").

Multi_organ_candidate_cpg <- methylation_data_list$Multi_organ$Probe
Brain_candidate_cpg     <- methylation_data_list$Brain$Probe
Caudal_Fin_candidate_cpg       <- methylation_data_list$Caudal_Fin$Probe
Intestines_candidate_cpg <- methylation_data_list$Intestines$Probe

Multi_organ_selected_cpg  <- Multi_organ_best_cpg
Brain_selected_cpg      <-Multi_organ_best_cpg
Caudal_Fin_selected_cpg        <-Multi_organ_best_cpg
Intestines_selected_cpg  <-Multi_organ_best_cpg


# need to make sure cpgs match annotation_database 

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}

Multi_organ_candidate_cpg <-adding_chr_cpg(Multi_organ_candidate_cpg)
Brain_candidate_cpg     <-adding_chr_cpg(Brain_candidate_cpg  )
Caudal_Fin_candidate_cpg       <-adding_chr_cpg(Caudal_Fin_candidate_cpg   )
Intestines_candidate_cpg <-adding_chr_cpg(Intestines_candidate_cpg)

Multi_organ_selected_cpg  <- adding_chr_cpg(Multi_organ_selected_cpg)
Brain_selected_cpg      <-adding_chr_cpg(Brain_selected_cpg )
Caudal_Fin_selected_cpg <- adding_chr_cpg(Caudal_Fin_selected_cpg)
Intestines_selected_cpg  <-adding_chr_cpg(Intestines_selected_cpg)


#--------------------------------------------------------------
# 4. parse selected cpgs 
#--------------------------------------------------------------
organs <- c("Multi_organ", "Brain", "Caudal_Fin" , "Intestines")
organ_list <- list ()

for (organ in organs) {
 
  message("Starting organ :", organ)
  
      selected_cpg <- get(paste0(organ, "_selected_cpg"))
    candidate_cpg <- get(paste0(organ, "_candidate_cpg"))
  
  # Function to ensure "chr" prefix
ensure_chr_prefix <- function(cpg_vector) {
  # Only add 'chr' if not already present
  cpg_vector <- ifelse(grepl("^chr", cpg_vector), cpg_vector, paste0("chr", cpg_vector))
  return(cpg_vector)
}

selected_cpg  <- ensure_chr_prefix(selected_cpg)
candidate_cpg <- ensure_chr_prefix(candidate_cpg)

  #----------------------------------------------------------
  # Parse genomic coordinates
  #----------------------------------------------------------
  selected_df <- data.frame(cpg = selected_cpg, stringsAsFactors = FALSE) %>%
    tidyr::separate(cpg, into = c("chr","start","end"), sep=":", fill="right") %>%
    mutate(start = as.numeric(start), end = as.numeric(end))
  
  candidate_df <- data.frame(cpg = candidate_cpg, stringsAsFactors = FALSE) %>%
    tidyr::separate(cpg, into = c("chr","start","end"), sep=":", fill="right") %>%
    mutate(start = as.numeric(start), end = as.numeric(end))
  
  #----------------------------------------------------------
  # Map to nearest genes using GRanges
  #----------------------------------------------------------
  selected_gr <- GRanges(seqnames = selected_df$chr,
                         ranges = IRanges(start=selected_df$start, end=selected_df$end))
  candidate_gr <- GRanges(seqnames = candidate_df$chr,
                          ranges = IRanges(start=candidate_df$start, end=candidate_df$end))
  
  genes_gr <- genes(TxDb.Drerio.UCSC.danRer11.refGene)
  selected_df$entrez  <- names(genes_gr)[nearest(selected_gr, genes_gr)]
  candidate_df$entrez <- names(genes_gr)[nearest(candidate_gr, genes_gr)]
  
  # Store per organ
  organ_list[[organ]] <- list(
    selected_cpg_df = selected_df,
    candidate_cpg_df = candidate_df
  )
}

#--------------------------------------------------------------
# 5. Convert Human PCGT Genes to Zebrafish Entrez IDs
#--------------------------------------------------------------
human <- useEnsembl("ensembl", dataset="hsapiens_gene_ensembl", mirror="useast")
pcgt_genes_human <- unique(na.omit(Bracken2006_PCGT_human$Gene))

# Get zebrafish orthologue Ensembl IDs
ortholog_table <- getBM(
  attributes=c("drerio_homolog_ensembl_gene","drerio_homolog_associated_gene_name"),
  filters="hgnc_symbol",
  values=pcgt_genes_human,
  mart=human
)

zfish_ensembl <- unique(ortholog_table$drerio_homolog_ensembl_gene)

# Map zebrafish Ensembl IDs to Entrez IDs
zebrafish <- useEnsembl("ensembl", dataset="drerio_gene_ensembl", mirror="useast")
zfish_mapping <- getBM(
  attributes=c("ensembl_gene_id","entrezgene_id"),
  filters="ensembl_gene_id",
  values=zfish_ensembl,
  mart=zebrafish
)

pcgt_zfish_entrez <- unique(na.omit(zfish_mapping$entrezgene_id))

#--------------------------------------------------------------
# 6. Compute PCGT Enrichment per Organ
#--------------------------------------------------------------
enrichment_results <- lapply(names(organ_list), function(org) {
  message("Processing organ: ", org)
  
  tryCatch({
    

  selected_df  <- organ_list[[org]]$selected_cpg_df
  candidate_df <- organ_list[[org]]$candidate_cpg_df
  
  all_genes <- unique(c(candidate_df$entrez, selected_df$entrez))
  pcgt_flag <- all_genes %in% pcgt_zfish_entrez
  clock_flag <- all_genes %in% selected_df$entrez
  
  contingency <- table(pcgt_flag, clock_flag)
  fisher_res <- fisher.test(contingency)
  
  data.frame(
    organ = org,
    n_selected = nrow(selected_df),
    n_candidate = nrow(candidate_df),
    overlap_n = sum(selected_df$entrez %in% pcgt_zfish_entrez),
    overlap_prop = mean(selected_df$entrez %in% pcgt_zfish_entrez),
    p_value = fisher_res$p.value,
    odds_ratio = fisher_res$estimate
  )
  } , error = function(e) {
    message("Processing failed for ", organ, ": ", e$message)
    })
  
})

enrichment_df <- do.call(rbind, enrichment_results)

#--------------------------------------------------------------
# 7. Plot PCGT Enrichment
#--------------------------------------------------------------
# Create significance labels

enrichment_df <- enrichment_df %>%
  mutate(sig_label = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    p_value < 0.10  ~ ".",
    TRUE            ~ ""
  ))

# Compute standard error for error bars
enrichment_df <- enrichment_df %>%
  rowwise() %>%
  mutate(
    se = sqrt(overlap_prop * (1 - overlap_prop) / n_selected),
    ymin = overlap_prop - se,
    ymax = overlap_prop + se,
    sig_y = ymax + 0.05,  # y-position for significance label above error bar,
    organ_plot = "Soma"
  ) 

# Plot with significance labels
p <- ggplot(enrichment_df, aes(x=organ_plot, y=overlap_prop, color=organ_plot)) +
  geom_point(size=4) +
  geom_errorbar(aes(ymin=ymin, ymax=ymax), width=0.2) +
  geom_label(aes(label=sprintf("%.2f", overlap_prop)),
             vjust=-0.5, size=4, fill="white", color="black") +
  geom_text(aes(y=sig_y, label=sig_label), vjust=0, size=5, color="black") +
  labs(title="PCGT Orthologue Enrichment per Organ",
       y="Proportion of CpGs near PCGT orthologues",
       x=NULL) +
  scale_color_manual(values=okabe_ito_palette(1)) +
  
  theme(legend.position="none")  +
  plot_theme 

print(p)

ggsave(
  filename = file.path(output_directory, "PCGT_genes_universal_Multi_organ.png"),
  plot = p,
  height = 7,
  width = 7,
  dpi = 400
)

```

### Universal overlapping soma
```{r bb}
##############################################################
# Zebrafish CpG Model Analysis & PCGT Enrichment Pipeline
# Description: Select best CpG-based epigenetic clock models 
# per organ, map CpGs to genes, convert human PCGT genes to 
# zebrafish orthologues, and test enrichment.
##############################################################

#--------------------------------------------------------------
# 0. Load Required Libraries
#--------------------------------------------------------------
suppressPackageStartupMessages({
library(dplyr)                  
library(tidyr)                  
library(GenomicRanges)          
library(TxDb.Drerio.UCSC.danRer11.refGene) 
library(clusterProfiler)        
library(org.Dr.eg.db)           
library(missMethyl)             
library(gprofiler2)             
library(biomaRt)                
library(ggplot2)
  library(readr)
})
#--------------------------------------------------------------
# 1. Load Data
#--------------------------------------------------------------
# merged_data$metrics: performance metrics (MAE, R²) per model
# cpg_data_validation: CpGs per model
# best_model_*: fitted model objects per organ
# Bracken2006_PCGT_human: human PCGT gene list


Bracken2006_PCGT_human <- read_csv(here("data/Bracken_et_al_2006_PCGT_list/Bracken2006_PCCT_human.csv"))

#--------------------------------------------------------------
# 3. Extract CpGs from Best Models
#--------------------------------------------------------------

Multi_organ_best_cpg  <- unique(best_cpg_overlap_single)


meta_data <- list(
  Multi_organ      = best_model_list$Multi_organ$sample_data_filtered,
  Brain           = best_model_list$Brain$sample_data_filtered,
  Caudal_Fin = best_model_list$Caudal_Fin$sample_data_filtered,
  Intestines      = best_model_list$Intestines$sample_data_filtered
)


#--------------------------------------------------------------
# 5. Extract Candidate and Selected CpGs
#--------------------------------------------------------------
# Candidate CpGs = all CpGs available for model training.
# Selected CpGs = CpGs retained in the final best model.
# Both are represented as genomic coordinate strings 
# (e.g., "chr1:24557370:24557869").

Brain_candidate_cpg     <- methylation_data_list$Brain$Probe
Caudal_Fin_candidate_cpg       <- methylation_data_list$Caudal_Fin$Probe
Intestines_candidate_cpg <- methylation_data_list$Intestines$Probe


Brain_selected_cpg      <-Multi_organ_best_cpg
Caudal_Fin_selected_cpg        <-Multi_organ_best_cpg
Intestines_selected_cpg  <-Multi_organ_best_cpg


# need to make sure cpgs match annotation_database 

adding_chr_cpg <- function(df){
   df <- sub("^(\\d+)", "chr\\1", df)
    df
}


Brain_candidate_cpg     <-adding_chr_cpg(Brain_candidate_cpg  )
Caudal_Fin_candidate_cpg       <-adding_chr_cpg(Caudal_Fin_candidate_cpg   )
Intestines_candidate_cpg <-adding_chr_cpg(Intestines_candidate_cpg)

Brain_selected_cpg      <-adding_chr_cpg(Brain_selected_cpg )
Caudal_Fin_selected_cpg <- adding_chr_cpg(Caudal_Fin_selected_cpg)
Intestines_selected_cpg  <-adding_chr_cpg(Intestines_selected_cpg)


#--------------------------------------------------------------
# 4. parse selected cpgs 
#--------------------------------------------------------------
organs <- c( "Brain", "Caudal_Fin" , "Intestines")
organ_list <- list ()

for (organ in organs) {
 
  message("Starting organ :", organ)
  
      selected_cpg <- get(paste0(organ, "_selected_cpg"))
    candidate_cpg <- get(paste0(organ, "_candidate_cpg"))
  
  # Function to ensure "chr" prefix
ensure_chr_prefix <- function(cpg_vector) {
  # Only add 'chr' if not already present
  cpg_vector <- ifelse(grepl("^chr", cpg_vector), cpg_vector, paste0("chr", cpg_vector))
  return(cpg_vector)
}

selected_cpg  <- ensure_chr_prefix(selected_cpg)
candidate_cpg <- ensure_chr_prefix(candidate_cpg)

  #----------------------------------------------------------
  # Parse genomic coordinates
  #----------------------------------------------------------
  selected_df <- data.frame(cpg = selected_cpg, stringsAsFactors = FALSE) %>%
    tidyr::separate(cpg, into = c("chr","start","end"), sep=":", fill="right") %>%
    mutate(start = as.numeric(start), end = as.numeric(end))
  
  candidate_df <- data.frame(cpg = candidate_cpg, stringsAsFactors = FALSE) %>%
    tidyr::separate(cpg, into = c("chr","start","end"), sep=":", fill="right") %>%
    mutate(start = as.numeric(start), end = as.numeric(end))
  
  #----------------------------------------------------------
  # Map to nearest genes using GRanges
  #----------------------------------------------------------
  selected_gr <- GRanges(seqnames = selected_df$chr,
                         ranges = IRanges(start=selected_df$start, end=selected_df$end))
  candidate_gr <- GRanges(seqnames = candidate_df$chr,
                          ranges = IRanges(start=candidate_df$start, end=candidate_df$end))
  
  genes_gr <- genes(TxDb.Drerio.UCSC.danRer11.refGene)
  selected_df$entrez  <- names(genes_gr)[nearest(selected_gr, genes_gr)]
  candidate_df$entrez <- names(genes_gr)[nearest(candidate_gr, genes_gr)]
  
  # Store per organ
  organ_list[[organ]] <- list(
    selected_cpg_df = selected_df,
    candidate_cpg_df = candidate_df
  )
}

#--------------------------------------------------------------
# 5. Convert Human PCGT Genes to Zebrafish Entrez IDs
#--------------------------------------------------------------
human <- useEnsembl("ensembl", dataset="hsapiens_gene_ensembl", mirror="useast")
pcgt_genes_human <- unique(na.omit(Bracken2006_PCGT_human$Gene))

# Get zebrafish orthologue Ensembl IDs
ortholog_table <- getBM(
  attributes=c("drerio_homolog_ensembl_gene","drerio_homolog_associated_gene_name"),
  filters="hgnc_symbol",
  values=pcgt_genes_human,
  mart=human
)

zfish_ensembl <- unique(ortholog_table$drerio_homolog_ensembl_gene)

# Map zebrafish Ensembl IDs to Entrez IDs
zebrafish <- useEnsembl("ensembl", dataset="drerio_gene_ensembl", mirror="useast")
zfish_mapping <- getBM(
  attributes=c("ensembl_gene_id","entrezgene_id"),
  filters="ensembl_gene_id",
  values=zfish_ensembl,
  mart=zebrafish
)

pcgt_zfish_entrez <- unique(na.omit(zfish_mapping$entrezgene_id))

#--------------------------------------------------------------
# 6. Compute PCGT Enrichment per Organ
#--------------------------------------------------------------
enrichment_results <- lapply(names(organ_list), function(org) {
  message("Processing organ: ", org)
  
  tryCatch({
    

  selected_df  <- organ_list[[org]]$selected_cpg_df
  candidate_df <- organ_list[[org]]$candidate_cpg_df
  
  all_genes <- unique(c(candidate_df$entrez, selected_df$entrez))
  pcgt_flag <- all_genes %in% pcgt_zfish_entrez
  clock_flag <- all_genes %in% selected_df$entrez
  
  contingency <- table(pcgt_flag, clock_flag)
  fisher_res <- fisher.test(contingency)
  
  data.frame(
    organ = org,
    n_selected = nrow(selected_df),
    n_candidate = nrow(candidate_df),
    overlap_n = sum(selected_df$entrez %in% pcgt_zfish_entrez),
    overlap_prop = mean(selected_df$entrez %in% pcgt_zfish_entrez),
    p_value = fisher_res$p.value,
    odds_ratio = fisher_res$estimate
  )
  } , error = function(e) {
    message("Processing failed for ", organ, ": ", e$message)
    })
  
})

enrichment_df <- do.call(rbind, enrichment_results)

#--------------------------------------------------------------
# 7. Plot PCGT Enrichment
#--------------------------------------------------------------
# Create significance labels

enrichment_df <- enrichment_df %>%
  mutate(sig_label = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    p_value < 0.10  ~ ".",
    TRUE            ~ ""
  ))

# Compute standard error for error bars
enrichment_df <- enrichment_df %>%
  rowwise() %>%
  mutate(
    se = sqrt(overlap_prop * (1 - overlap_prop) / n_selected),
    ymin = overlap_prop - se,
    ymax = overlap_prop + se,
    sig_y = ymax + 0.05,  # y-position for significance label above error bar,
    organ_plot = "Soma"
  ) 

# Plot with significance labels
p <- ggplot(enrichment_df, aes(x=organ_plot, y=overlap_prop, color=organ_plot)) +
  geom_point(size=4) +
  geom_errorbar(aes(ymin=ymin, ymax=ymax), width=0.2) +
  geom_label(aes(label=sprintf("%.2f", overlap_prop)),
             vjust=-0.5, size=4, fill="white", color="black") +
  geom_text(aes(y=sig_y, label=sig_label), vjust=0, size=5, color="black") +
  labs(title="PCGT Orthologue Enrichment per Organ",
       y="Proportion of CpGs near PCGT orthologues", colour= "",
       x=NULL) +
  scale_color_manual(values=okabe_ito_palette(1)) +
  guides(colour="none")+
  theme(legend.position="none")  +
  plot_theme 

print(p)
 pcgt_overlap_soma <- p

ggsave(
  filename = file.path(output_directory, "PCGT_genes_universal_soma_overlap.png"),
  plot =  pcgt_overlap_soma,
  height = 7,
  width = 7,
  dpi = 400
)

```

# ~ 

## Quantify directionality of methylation drift (age-related hyper/hypo)

To test whether CpGs generally move toward intermediate methylation (stochastic drift) or follow Polycomb-like hypermethylation at promoters.

### Organ specific 
```{r cc, message=FALSE}

# ============================================================
# Overlap-based mapping + drift pipeline
# ============================================================

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(data.table)
})

# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
orgs <- c("Multi_organ", "Brain", "Caudal_Fin" , "Intestines")

selected_cpg <- best_selected_cpgs_organs

# ============================================================
# 2. filter methylaiton by best Cpgs 
# ============================================================


meth_list <- list()
for(o in orgs){
  cpg <- selected_cpg[[o]]
  
  
  df <- methylation_best_cpgs %>%
  filter(tile%in%  cpg)
  
  if(o !="Multi_organ"){
    df <- df %>%
      filter(Organ == o)
  }
  
   df$organ_value <- o
  
  
  meth_list[[o]] <- df
}
#============================================================
# 3. Rbind everything together into one dataframe 
#------------------------------------------------------------


merged_meth <- bind_rows(meth_list )

merged_meth <- merged_meth %>%
  mutate(organ_value = factor(organ_value, levels = c("Brain","Caudal_Fin" ,"Intestines","Multi_organ")
  ),
                        meth=methylation,
  CpG = tile
  )

slopes_all <- merged_meth %>%
  group_by(Organ, CpG) %>%
  summarise(
    slope     = coef(lm(meth ~ Age))[2],
    mean_meth = median(meth, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(deviation = mean_meth - 0.5)

cor_deviation_slope <- slopes_all %>%
  group_by(Organ) %>%
  summarise(cor_dev_slope = cor(deviation, slope, use = "complete.obs"))

```

####  Slope vs mean methylation
```{r dd}

p <- ggplot(slopes_all, aes(x = mean_meth, y = slope, color = Organ, fill = Organ)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Mean methylation",
    y = "dM/dAge",
    title = "CpG methylation drift across organs",
    colour = ""
  ) +
  guides(fill = "none")+
  scale_colour_manual(values = master_pal)+
  scale_fill_manual(values = master_pal)+

  theme(legend.position = "bottom")  +
  plot_theme 

p 
methylation_drift_org_specific <-p 
ggsave(
  filename = paste0(output_directory, "/Methylation_drift_all_cpg_organ_specific.png"),
  plot = methylation_drift_org_specific,
  height = 5,
  width = 7,
  dpi = 400
)


```

####  Density slope plot
```{r ee}
# quantify whether organs have a bias toward hyper- or hypomethylation.

m<- lmer(slope ~ Organ +(1|CpG), data = slopes_all)

summary(m)
Anova(m, type = "III")
emmeans(m,pairwise ~Organ)

p<- ggplot(slopes_all, aes(x = slope, fill = Organ, colour = Organ)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "dM/dAge", y = "Density", title = "Distribution of CpG methylation slopes",fill = "") +
  scale_fill_manual(values = master_pal) +
  scale_colour_manual(values = master_pal) +
  guides(colour = "none")+
  theme_classic2(base_size = 16)+
  theme(legend.position = "bottom")  +
  plot_theme 
p 
methylation_density_org_specific <- p
ggsave(
  filename = paste0(output_directory, "/Density_slope_organ_specific.png"),
  plot = methylation_density_org_specific,
  height = 7,
  width = 6,
  dpi = 400
)
```

#### Correlation heatmap of deviation vs slope
```{r ff}
#concise summary of which organs show a drift toward intermediate methylation
# bootstrap correlation per organ
set.seed(42)
nboot <- 1000

boot_cor <- slopes_all %>%
  group_by(Organ) %>%
  summarise(
    cor_boot = list({
      sapply(1:nboot, function(i) {
        samp <- sample(n(), replace = TRUE)
        cor(deviation[samp], slope[samp], use = "complete.obs")
      })
    }),
    .groups = "drop"
  )

# summarise mean ± SE
boot_summary <- boot_cor %>%
  rowwise() %>%
  mutate(
    mean_cor = mean(unlist(cor_boot)),
    se_cor   = sd(unlist(cor_boot)) / sqrt(length(unlist(cor_boot)))
  )

# plot with error bars
p <- ggplot(boot_summary, aes(x = Organ, y = mean_cor, fill = Organ)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_cor - se_cor, ymax = mean_cor + se_cor), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "Correlation (deviation vs slope)", title = "Stochastic drift signature per organ", fill = "", x = "") +
  scale_fill_manual(values = master_pal) +

  theme(legend.position = "bottom")  +
  plot_theme 

p
stochastic_drift_org_specific <- p
ggsave(
  filename = paste0(output_directory, "/Deviation_slope_heatmap_organ_specific.png"),
  plot = stochastic_drift_org_specific,
  height = 7,
  width = 6,
  dpi = 400
)

# linear model with interaction
mod <- lmer(slope ~ deviation * Organ + (1|CpG), data = slopes_all)
summary(mod)
Anova(mod, type="III")
emtr <- emtrends(
  mod,
  ~ Organ,
  var = "deviation"
)

emtr
pairs(emtr)
```

### Universal all organs 
```{r gg, message=FALSE}

# ============================================================
# Overlap-based mapping + drift pipeline
# ============================================================

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
})
# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
orgs <- c("Multi_organ", "Brain", "Caudal_Fin" , "Intestines")

selected_cpg <- best_cpg_overlap_all

# ============================================================
# 2. filter methylaiton by best Cpgs 
# ============================================================


  df <- methylation_best_cpgs %>%
  filter(tile%in%  selected_cpg)
  

#============================================================
# 3. Rbind everything together into one dataframe 
#------------------------------------------------------------


merged_meth <- df

merged_meth <- merged_meth %>%
  mutate(Organ= factor(Organ, levels = c("Brain","Caudal_Fin" ,"Intestines")
  ),
                        meth=methylation,
  CpG = tile
  )
slopes_all <- merged_meth %>%
  group_by(Organ, CpG) %>%
  summarise(
    slope     = coef(lm(meth ~ Age))[2],
    mean_meth = median(meth, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(deviation = mean_meth - 0.5)

cor_deviation_slope <- slopes_all %>%
  group_by(Organ) %>%
  summarise(cor_dev_slope = cor(deviation, slope, use = "complete.obs"))

```

####  Slope vs mean methylation
```{r hh}

p <- ggplot(slopes_all, aes(x = mean_meth, y = slope, color = Organ, fill = Organ)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Mean methylation",
    y = "dM/dAge",
    title = "CpG methylation drift across organs",
    colour = ""
  ) +
  guides(fill = "none")+
  scale_colour_manual(values = master_pal)+
  scale_fill_manual(values = master_pal)+
  
  plot_theme 

p 

ggsave(
  filename = paste0(output_directory, "/Methylation_drift_all_cpg_universal_all_models.png"),
  plot = p,
  height = 5,
  width = 8,
  dpi = 400
)


```

####  Density slope plot
```{r ii}
# quantify whether organs have a bias toward hyper- or hypomethylation.

m<- lmer(slope ~ Organ +(1|CpG), data = slopes_all)

summary(m)
Anova(m, type = "III")
emmeans(m,pairwise ~Organ)

p<- ggplot(slopes_all, aes(x = slope, fill = Organ, colour = Organ)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "dM/dAge", y = "Density", title = "Distribution of CpG methylation slopes",fill = "") +
  scale_fill_manual(values = master_pal) +
  scale_colour_manual(values = master_pal) +
  guides(colour = "none")+
  theme_classic2(base_size = 16)+
  theme(legend.position = "bottom")  +
  plot_theme 
p 

ggsave(
  filename = paste0(output_directory, "/Density_slope_universal_all_models.png"),
  plot = p,
  height = 5,
  width = 7,
  dpi = 400
)
```

#### Correlation heatmap of deviation vs slope
```{r jj}
#concise summary of which organs show a drift toward intermediate methylation
# bootstrap correlation per organ
set.seed(42)
nboot <- 1000

boot_cor <- slopes_all %>%
  group_by(Organ) %>%
  summarise(
    cor_boot = list({
      sapply(1:nboot, function(i) {
        samp <- sample(n(), replace = TRUE)
        cor(deviation[samp], slope[samp], use = "complete.obs")
      })
    }),
    .groups = "drop"
  )

# summarise mean ± SE
boot_summary <- boot_cor %>%
  rowwise() %>%
  mutate(
    mean_cor = mean(unlist(cor_boot)),
    se_cor   = sd(unlist(cor_boot)) / sqrt(length(unlist(cor_boot)))
  )

# plot with error bars
p <- ggplot(boot_summary, aes(x = Organ, y = mean_cor, fill = Organ)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_cor - se_cor, ymax = mean_cor + se_cor), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "Correlation (deviation vs slope)", title = "Stochastic drift signature per organ", fill = "") +
  scale_fill_manual(values = master_pal) +
  
  plot_theme 

p

ggsave(
  filename = paste0(output_directory, "/Deviation_slope_heatmap_universal_all_models.png"),
  plot = p,
  height = 5,
  width = 8,
  dpi = 400
)




# linear model with interaction
mod <- lmer(slope ~ deviation * Organ + (1|CpG), data = slopes_all)
summary(mod)
Anova(mod, type="III")
emtr <- emtrends(
  mod,
  ~ Organ,
  var = "deviation"
)

emtr
pairs(emtr)
```


### Universal overlappign soma
```{r kk, message=FALSE}

# ============================================================
# Overlap-based mapping + drift pipeline
# ============================================================

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
})

# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
orgs <- c( "Brain", "Caudal_Fin" , "Intestines")

selected_cpg <- best_cpg_overlap_single

# ============================================================
# 2. filter methylaiton by best Cpgs 
# ============================================================

  df <- methylation_best_cpgs %>%
  filter(tile%in%  selected_cpg)
  

#============================================================
# 3. Rbind everything together into one dataframe 
#------------------------------------------------------------


merged_meth <- df

merged_meth <- merged_meth %>%
  mutate(Organ = factor(Organ, levels = c("Brain","Caudal_Fin" ,"Intestines","Multi_organ")
  ),
                        meth=methylation,
  CpG = tile 
  )

slopes_all <- merged_meth %>%
  group_by(Organ, CpG) %>%
  summarise(
    slope     = coef(lm(meth ~ Age))[2],
    mean_meth = median(meth, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(deviation = mean_meth - 0.5)

cor_deviation_slope <- slopes_all %>%
  group_by(Organ) %>%
  summarise(cor_dev_slope = cor(deviation, slope, use = "complete.obs"))

```

####  Slope vs mean methylation
```{r ll}
p <- ggplot(slopes_all, aes(x = mean_meth, y = slope, color = Organ, fill = Organ)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Mean methylation",
    y = "dM/dAge",
    title = "CpG methylation drift across organs",
    colour = ""
  ) +
  guides(fill = "none")+
  scale_colour_manual(values = master_pal)+
  scale_fill_manual(values = master_pal)+
  
  plot_theme 

p 
methylation_drift_overlap_soma <- p
ggsave(
  filename = paste0(output_directory, "/Methylation_drift_all_cpg_universal_soma_overlap.png"),
  plot = methylation_drift_overlap_soma,
  height = 7,
  width = 6,
  dpi = 400
)

```

####  Density slope plot
```{r mm}
# quantify whether organs have a bias toward hyper- or hypomethylation.

m<- lmer(slope ~ Organ +(1|CpG), data = slopes_all)

summary(m)
Anova(m, type = "III")
emmeans(m,pairwise ~Organ)

p<- ggplot(slopes_all, aes(x = slope, fill = Organ, colour = Organ)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "dM/dAge", y = "Density", title = "Distribution of CpG methylation slopes",fill = "") +
  scale_fill_manual(values = master_pal) +
  scale_colour_manual(values = master_pal) +
  guides(colour = "none")+
  theme_classic2(base_size = 16)+
  theme(legend.position = "bottom")  +
  plot_theme 
p 
methylation_density_overlap_soma <- p

ggsave(
  filename = paste0(output_directory, "/Density_slope_universal_soma_overlap.png"),
  plot = methylation_density_overlap_soma,
  height = 7,
  width = 6,
  dpi = 400
)
```

#### Correlation heatmap of deviation vs slope
```{r nn}
# Correlation heatmap of deviation vs slope
#concise summary of which organs show a drift toward intermediate methylation
# bootstrap correlation per organ
set.seed(42)
nboot <- 1000

boot_cor <- slopes_all %>%
  group_by(Organ) %>%
  summarise(
    cor_boot = list({
      sapply(1:nboot, function(i) {
        samp <- sample(n(), replace = TRUE)
        cor(deviation[samp], slope[samp], use = "complete.obs")
      })
    }),
    .groups = "drop"
  )

# summarise mean ± SE
boot_summary <- boot_cor %>%
  rowwise() %>%
  mutate(
    mean_cor = mean(unlist(cor_boot)),
    se_cor   = sd(unlist(cor_boot)) / sqrt(length(unlist(cor_boot)))
  )

# plot with error bars
p <- ggplot(boot_summary, aes(x = Organ, y = mean_cor, fill = Organ)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_cor - se_cor, ymax = mean_cor + se_cor), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "Correlation (deviation vs slope)", title = "Stochastic drift signature per organ", fill = "", x = "") +
  scale_fill_manual(values = master_pal) +
  
  plot_theme 

p
stochastic_drift_overlap_soma <- p

ggsave(
  filename = paste0(output_directory, "/Deviation_slope_heatmap_universal_soma_overlap.png"),
  plot = stochastic_drift_overlap_soma,
  height = 7,
  width = 6,
  dpi = 400
)


# linear model with interaction
mod <- lmer(slope ~ deviation * Organ + (1|CpG), data = slopes_all)
summary(mod)
Anova(mod, type="III")
emtr <- emtrends(
  mod,
  ~ Organ,
  var = "deviation"
)

emtr
pairs(emtr)
```
### ~

