---
title: "Cross-Tissue analysis - part 8: dynamic_network_pca_entropy"
author: "Jean-Charles de Coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
PIPELINE STAGE: 06 / Aim 2: Cross-Tissue Ageing Biology - part 8 of 10
=================================================================
Tile-level and tissue-level PCA of methylation profiles, plus methylation variance and entropy analysis per tissue.

Depends on objects created by part 7 (07_methylation_drift_and_pcgt_enrichment.Rmd) having already been run in the same R
session (or knitted first, with its environment preserved) - this file does not reload raw data from scratch.
=================================================================
-->

```{r config_part8}
# Set the root directory of this project (edit here, or export
# EPICLOCK_ROOT before running).
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
```

# Dynamic and network-level analyses

PCA or network clustering on all model CpGs to identify organ groupings and systemic mehtylation modules 

Entropy or variane of methyaltion patters per orga to quantify stochastic epigenetic drift - https://doi.org/10.1111/acel.13492

##Combined network or PCA for organ grouping

To detect shared methylation signatures across organs:
```{r oo, message=FALSE, warning=FALSE }
#============================================================
# Libraries
#------------------------------------------------------------
suppressPackageStartupMessages({
library(dplyr)        # Data wrangling
library(ggplot2)      # Plotting
library(ggfortify)    # autoplot for PCA
library(matrixStats)  # row/column summaries
library(ggpubr)     # theme_classic2 (optional)
})

# ============================================================
# 1. Define organ list and CpG selections
# ============================================================
# Specify the organs to include in the analysis
orgs <- c("Brain", "Caudal_Fin", "Intestines")

# Select the CpGs of interest for PCA (typically robust or top-ranked)
selected_cpg <- best_cpg_overlap_single


# ============================================================
# 2. Filter methylation data by selected CpGs
# ============================================================
# Keep only the methylation measurements corresponding to the selected CpGs
df <- methylation_best_cpgs %>%
  filter(tile %in% selected_cpg)


# ============================================================
# 3. Prepare merged methylation dataframe
# ============================================================
# In this case, df already contains all selected CpGs, so we assign it directly
merged_meth <- df

# Select only the columns required for PCA: tile ID, methylation value, and organ
merged_meth_matrix <- merged_meth %>%
  dplyr::select(tile, methylation, Organ)


# ============================================================
# 4. Summarise methylation per tile × organ and create pca file
# ============================================================
# Compute the mean methylation per tile in each organ
# This collapses any replicate measurements (technical or biological) into a single value
tile_organ_means <- merged_meth_matrix %>%
  group_by(tile, Organ) %>%
  summarise(
    methylation = mean(methylation, na.rm = TRUE),
    .groups = "drop"
  )
tile_mat <- tile_organ_means %>%
    pivot_wider(names_from = Organ, values_from = methylation) %>%
    column_to_rownames("tile") %>%
    as.matrix()

tile_sd <- apply(tile_mat, 1, sd, na.rm = TRUE)
tile_mat_filt <- tile_mat[tile_sd > 0 & !is.na(tile_sd), ]


pca_tile <- prcomp(tile_mat_filt, scale. = TRUE)

# ============================================================
# 5. Determine "dominant organ" for each tile
# ============================================================
# For visualization purposes, assign each tile to the organ in which its methylation is highest
# This is used to colour points in the tile-centric PCA
tile_labels <- tile_organ_means %>%
  group_by(tile) %>%
  slice_max(methylation, n = 1, with_ties = FALSE) %>%  # choose one organ per tile
  dplyr::select(tile, dominant_organ = Organ)



# ============================================================
# 6. Tile-centric PCA plot
# ============================================================
# Tiles are the observations (rows), organs are the features (columns)
# PCA reveals patterns of organ-specific methylation across tiles
p <- autoplot(
  pca_tile,                # PCA object for tile-centric analysis
  data = tile_labels,      # colour tiles by their "dominant organ"
  colour = "dominant_organ",
  alpha = 0.6,
  size = 2
) +

  scale_colour_manual(values = okabe_ito_palette(3)) +
  labs(
    subtitle = "Tiles coloured by organ of maximal methylation",
    colour = "Dominant Organ"
  )+
  theme(legend.position = "bottom")  +
  plot_theme 

p
# Save figure
ggsave(
  filename = file.path(output_directory, "Tile_PCA_plot_overlapping_Soma.png"),
  plot = p,
  height = 5,
  width = 7,
  dpi = 400
)
# ============================================================
# 7. Organ-centric PCA
# ============================================================
# Prepare a PCA where organs are the observations
# Start with the same filtered tile × organ matrix
# Transpose so rows = organs, columns = tiles
organ_mat <- t(tile_mat_filt)

# Remove any tiles that have zero variance across organs
# This avoids errors in scaling and PCA computation
tile_sd2 <- apply(organ_mat, 2, sd, na.rm = TRUE)
organ_mat_filt <- organ_mat[, tile_sd2 > 0 & !is.na(tile_sd2)]

# Run PCA on organs (rows)
# Each organ is positioned in "tile methylation space"
pca_organ <- prcomp(organ_mat_filt, scale. = TRUE)

# Plot the organ-centric PCA
# Here, one point per organ is expected
# Loadings are omitted because tiles (features) vastly outnumber organs (observations)
p <- autoplot(
  pca_organ,
  data = data.frame(Organ = rownames(pca_organ$x)),
  colour = "Organ",
  size = 5
) +

  scale_colour_manual(values = okabe_ito_palette(3)) +
  labs(
title = "PCA",
    colour = ""
  )+
  theme(legend.position = "bottom")  +
  plot_theme 
p
pca_org_level <- p

# Save figure
ggsave(
  filename = file.path(output_directory, "Organ_PCA_plot_overlapping_Soma.png"),
  plot = pca_org_level,
  height = 5,
  width = 5,
  dpi = 400
)

```
 
## Methylation variance and entropy per organ

Entropy and variability indicate stochastic drift.

```{r pp}
suppressPackageStartupMessages({
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(ggfortify)
})
#============================================================
# 1️⃣ CpG-wise Shannon entropy per organ
#============================================================

selected_cpg <- best_cpg_overlap_single

df <- methylation_best_cpgs %>%
  filter(tile %in% selected_cpg)

merged_meth_matrix <- df %>%
  dplyr::select(tile, methylation, Organ, individual_fish_ID)


#  Summarise methylation per tile × organ
# Compute the mean methylation per tile in each organ
# This collapses any replicate measurements (technical or biological) into a single value
tile_organ_means <- merged_meth_matrix %>%
  group_by(tile, Organ, individual_fish_ID) %>%
  summarise(
    methylation = mean(methylation, na.rm = TRUE),
    .groups = "drop"
  )

entropy_per_cpg <- lapply(unique(tile_organ_means$Organ), function(org){
  
  # Subset df for this organ
  mat <- tile_organ_means %>% filter(Organ == org) %>%
    dplyr::select(tile, methylation, individual_fish_ID) %>%
    pivot_wider(names_from = individual_fish_ID, values_from = methylation)
  
  # Extract numeric matrix (rows = CpGs, cols = samples)
  mat_numeric <- as.matrix(mat[,-1]) / 100  # scale 0–1
  
  # clamp to avoid log2(0)
  mat_numeric <- pmin(pmax(mat_numeric, 1e-6), 1-1e-6)
  
  # Compute entropy per CpG across samples
  entropy <- apply(mat_numeric, 1, function(x){
    mean(-x*log2(x) - (1-x)*log2(1-x))
  })
  
  data.frame(
    organ = org,
    tile = mat$tile,
    entropy = entropy
  )
}) %>% bind_rows()


# Plot CpG-wise entropy
p_cpg <- ggplot(entropy_per_cpg, aes(x = organ, y = entropy, colour = organ, fill = organ)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  stat_summary(fun = mean, geom = "point", color = "black", show.legend = F) +
  stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width = 0.2, show.legend = F) +

  scale_colour_manual(values = okabe_ito_palette(length(unique(entropy_per_cpg$organ)))) +
  scale_fill_manual(values = okabe_ito_palette(length(unique(entropy_per_cpg$organ)))) +
  labs(title = "CpG-wise methylation entropy per organ",
       y = "Methylation heterogeneity (0–1)",
       x = "",
       colour = "",
       fill = "") +
  ylim(0,1) +
  theme(legend.position = "bottom")  +
  plot_theme 


signif_data <- get_signif_per_group(
  df = entropy_per_cpg,
  y_var = "entropy",
  x_var = "organ",
  group_var = "organ"
)
# Add dynamic significance results if they exist.
if (!is.null(signif_data) && nrow(signif_data) > 0) {
  signif_data$y.position <- signif_data$y.position

  p_cpg <- p_cpg + stat_pvalue_manual(
    signif_data,
    label = "label",
    tip.length = 0.0001,
    hide.ns = FALSE,
    size = 4
  )
}
p_cpg

ggsave(file.path(output_directory, "CpG_entropy_normalised_overlapping_soma.png"),
       plot = p_cpg, height = 7, width = 7, dpi = 400)


#============================================================
# 2️⃣ Sample-wise (global) entropy per organ
#============================================================

sample_entropy <- lapply(unique(tile_organ_means$Organ), function(org){
  
  mat <- tile_organ_means %>% filter(Organ == org) %>%
    dplyr::select(tile, methylation, individual_fish_ID) %>%
    pivot_wider(names_from = tile, values_from = methylation)
  
  mat_numeric <- as.matrix(mat[,-1]) / 100
  mat_numeric <- pmin(pmax(mat_numeric, 1e-6), 1-1e-6)
  
  # Row-wise entropy per sample
  entropy <- apply(mat_numeric, 1, function(x){
    mean(-x*log2(x) - (1-x)*log2(1-x))
  })
  
  data.frame(
    organ = org,
    sample_id = mat$individual_fish_ID,
    entropy = entropy
  )
}) %>% bind_rows()


# Plot sample-wise entropy
p_sample <- ggplot(sample_entropy, aes(x = organ, y = entropy, colour = organ, fill = organ)) +
  geom_violin(alpha = 0.6) +
  ggbeeswarm::geom_quasirandom(alpha = 0.2)+ 
  stat_summary(fun = mean, geom = "point", color = "black", show.legend = FALSE) +
  stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width = 0.2, show.legend = FALSE) +

  scale_colour_manual(values = okabe_ito_palette(length(unique(sample_entropy$organ)))) +
  scale_fill_manual(values = okabe_ito_palette(length(unique(sample_entropy$organ)))) +
  labs(title = "Sample-wise methylation entropy per organ",
       y = "Methylation heterogeneity (0–1)",
       x = " ",
       colour = "",
       fill = "") +
  plot_theme +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=20,hjust=1))  
  

signif_data <- get_signif_per_group(
  df = sample_entropy,
  y_var = "entropy",
  x_var = "organ",
  group_var = "organ",
    step_value = 7
)

# Add dynamic significance results if they exist.
if (!is.null(signif_data) && nrow(signif_data) > 0) {
  signif_data$y.position <- signif_data$y.position+ 0.05

  p_sample <- p_sample + stat_pvalue_manual(
    signif_data,
    label = "label",
    tip.length = 0.0001,
    hide.ns = FALSE,
    size = 4
  )
}

p_sample
entropy_plot <- p_sample

ggsave(file.path(output_directory, "Sample_entropy_normalised_overlapping_soma.png"),
       plot = entropy_plot, 
       height = 6, 
       width = 7, 
       dpi = 400)


m<- lmer(entropy ~ organ +(1|sample_id), data = sample_entropy)

summary(m)
Anova(m, type = "III")
emmeans(m,pairwise ~organ)
```

# ~ 
