---
title: "Check: CpG-set duplication and drift-direction"
author: "Jean-Charles de coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
SUPPLEMENTARY CHECK - not part of the main numbered pipeline (01-10).
=================================================================
Two questions in one script, because they share the same CpG-overlap
objects:

PART A - are the CpG sets behind Fig4D and Fig5D actually the same set
drawn twice (the two panels currently read as if they show independent
evidence)? Answered here using results_cpg_final.csv alone (tracked in
the repo - no excluded data needed).

PART B - re-derives the per-organ methylation-drift direction (slope vs
mean-methylation deviation) for each of the three CpG sets in question,
using methylation_cpgs.csv (also tracked; only the raw per-organ .tsv
matrices under data/Metrics_all_models/methylation_data/ are excluded).
This lets the sign/direction be compared directly against what Fig4D
states, addressing a figure-vs-text discrepancy.

This is fully self-contained - it does not require 02_quality_control_
and_model_matrices.Rmd or 03_best_model_ranking_significance.Rmd or
07_methylation_drift_and_pcgt_enrichment.Rmd to have been run first,
though it reuses their code verbatim.
=================================================================
-->

```{r config}
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(GenomicRanges)
  library(lme4)
  library(car)
  library(emmeans)
  library(ggplot2)
  library(readr)
})
```

# Load data (self-contained)

```{r load-data}
clean_up_data <- function(df) {
  df %>%
    filter(dataset == "Validation") %>%
    filter(!model %in% c("old_elastic_net", "old_lasso")) %>%
    mutate(
      rsq_threshold = dplyr::recode(rsq_threshold,
        "nofilter" = "r2_nofilter", "0.05" = "r2_0.05", "0.1" = "r2_0.1",
        "0.2" = "r2_0.2", "0.3" = "r2_0.3"),
      rsq_threshold = factor(rsq_threshold,
        levels = c("r2_nofilter","r2_0.05","r2_0.1","r2_0.2","r2_0.3")),
      filter = factor(filter, levels = c("nofilter","min5","min10","min20")),
      unique_model_id = paste(cpg_grouping, model_id, sep = "_")
    ) %>%
    mutate(organ = case_when(
      organ == "Fin_clip_caudal" ~ "Caudal_Fin",
      organ == "All_organs" ~ "Multi_organ",
      TRUE ~ organ
    ))
}

cpg_data <- read.csv(here("data/Metrics_all_models/final_merged/results_cpg_final.csv"))
cpg_data_validation <- clean_up_data(cpg_data)

cpg_data_validation_long <- cpg_data_validation %>%
  tidyr::separate_rows(selected_cpgs, sep = ";") %>%
  dplyr::mutate(selected_cpgs = trimws(selected_cpgs)) %>%
  mutate(selected_cpgs = gsub("chr", "", selected_cpgs)) %>%
  separate(selected_cpgs, c("chr", "start", "end"), remove = FALSE, sep = ":") %>%
  dplyr::mutate(chr = trimws(chr), chr = factor(chr, levels = c(1:25)))
```

# Part A: CpG-set overlap / duplication check

Verbatim reproduction of the overlap-computation logic from
`03_best_model_ranking_significance.Rmd`.

```{r overlap-computation}
fraction_threshold <- 0.05

gr_all <- GRanges(
  seqnames = cpg_data_validation_long$chr,
  ranges   = IRanges(start = as.numeric(cpg_data_validation_long$start),
                     end   = as.numeric(cpg_data_validation_long$end)),
  cpg      = cpg_data_validation_long$selected_cpgs,
  organ    = cpg_data_validation_long$organ,
  model_id = cpg_data_validation_long$unique_model_id
)

get_repeatable_regions <- function(gr, fraction_threshold = 0.05) {
  organs <- unique(gr$organ)
  repeatable_list <- map(organs, function(org_name) {
    message("Starting :", org_name)
    gr_org <- gr[gr$organ == org_name]
    n_runs <- length(unique(gr_org$model_id))
    threshold <- ceiling(fraction_threshold * n_runs)
    gr_reduced <- reduce(gr_org)
    counts <- sapply(seq_along(gr_reduced), function(i) {
      sum(countOverlaps(gr_org, gr_reduced[i]) > 0)
    })
    gr_repeatable <- gr_reduced[counts >= threshold]
    gr_repeatable$organ <- org_name
    gr_repeatable
  })
  names(repeatable_list) <- organs
  repeatable_list
}

repeatable_per_organ <- get_repeatable_regions(gr_all, fraction_threshold = fraction_threshold)

k1_df <- map_df(repeatable_per_organ, function(gr) {
  tibble(
    organs = gr$organ[1],
    chr = as.character(seqnames(gr)),
    start = start(gr),
    end = end(gr)
  )
}) %>%
  mutate(k = "k1", Probe = paste(chr, start, end, sep = ":")) %>%
  dplyr::select(k, organs, chr, start, end, Probe)

cpg_ranges <- cpg_data_validation_long %>%
  dplyr::select(cpg_grouping, organ, selected_cpgs, chr, start, end) %>%
  mutate(
    selected_cpgs = trimws(selected_cpgs),
    start = as.numeric(start),
    end   = as.numeric(end)
  )

cpg_granges_list <- cpg_ranges %>%
  group_by(organ) %>%
  summarise(
    gr = list(GRanges(seqnames = chr, ranges = IRanges(start = start, end = end), cpg = selected_cpgs)),
    .groups = "drop"
  ) %>%
  deframe()

organ_names <- names(cpg_granges_list)
overlap_results <- list()

for (k in 2:length(organ_names)) {
  combs <- combn(organ_names, k, simplify = FALSE)
  overlap_results[[paste0("k", k)]] <- map_df(combs, function(orgs_combo) {
    gr_list <- cpg_granges_list[orgs_combo]
    cpg_vectors <- lapply(gr_list, function(gr) gr$cpg)
    intersect_cpgs <- Reduce(intersect, cpg_vectors)
    if (length(intersect_cpgs) == 0) return(NULL)
    tibble(organs = paste(orgs_combo, collapse = ";"), cpg = intersect_cpgs)
  })
}

k2_plus_df <- bind_rows(overlap_results, .id = "k")
final_overlap_df <- bind_rows(k1_df, k2_plus_df)

best_cpg_overlap_all <- final_overlap_df %>%
  filter(k == "k4") %>%
  pull(cpg)

best_cpg_overlap_single <- final_overlap_df %>%
  filter(organs == "Brain;Caudal_Fin;Intestines") %>%
  pull(cpg)

best_selected_cpgs_organs <- list(
  Brain       = final_overlap_df %>% filter(organs == "Brain") %>% pull(Probe),
  Caudal_Fin  = final_overlap_df %>% filter(organs == "Caudal_Fin") %>% pull(Probe),
  Intestines  = final_overlap_df %>% filter(organs == "Intestines") %>% pull(Probe),
  Multi_organ = final_overlap_df %>% filter(organs == "Multi_organ") %>% pull(Probe)
)
```

```{r overlap-summary}
overlap_between_sets <- intersect(best_cpg_overlap_all, best_cpg_overlap_single)

cat("=== CpG set sizes ===\n")
cat("best_cpg_overlap_all    (Fig5D panel, k4 - all four organs):", length(best_cpg_overlap_all), "\n")
cat("best_cpg_overlap_single (Fig4D panel, Brain;Caudal_Fin;Intestines):", length(best_cpg_overlap_single), "\n")
cat("Manuscript states 30 shared loci for the relevant panel - compare against the counts above.\n\n")

cat("=== Direct overlap between the two panels' CpG sets ===\n")
cat("Shared CpGs:", length(overlap_between_sets), "\n")
cat("As % of best_cpg_overlap_all:", round(100 * length(overlap_between_sets) / length(best_cpg_overlap_all), 1), "%\n")
cat("As % of best_cpg_overlap_single:", round(100 * length(overlap_between_sets) / length(best_cpg_overlap_single), 1), "%\n\n")

if (length(overlap_between_sets) == length(best_cpg_overlap_all) ||
    length(overlap_between_sets) == length(best_cpg_overlap_single)) {
  cat("RESULT: one set is a full subset of the other - Fig4D and Fig5D are showing\n",
      "the same underlying CpGs (or a strict subset), not independent panels.\n")
} else if (length(overlap_between_sets) == 0) {
  cat("RESULT: no overlap - the two panels are drawing on genuinely disjoint CpG sets.\n")
} else {
  cat("RESULT: partial overlap - the two panels share some but not all CpGs.\n")
}
```

# Part B: drift-direction re-check for each CpG set

Verbatim reproduction of the `slopes_all` / `boot_cor` logic from
`07_methylation_drift_and_pcgt_enrichment.Rmd`, run once per CpG set so
the sign of the deviation-vs-slope correlation can be read off directly
and compared with what Fig4D currently states.

```{r load-methylation}
methylation_best_cpgs <- read_csv(here("data/Metrics_all_models/methylation_data/methylation_cpgs.csv"),
                                   show_col_types = FALSE) %>%
  mutate(
    Age = as.factor(Age),
    Organ = case_when(Organ == "Fin_clip_caudal" ~ "Caudal_Fin", TRUE ~ Organ)
  )
```

```{r drift-check-function}
run_drift_check <- function(selected_cpg, label, tissue_specific = FALSE) {

  orgs <- c("Multi_organ", "Brain", "Caudal_Fin", "Intestines")

  if (tissue_specific) {
    meth_list <- list()
    for (o in orgs) {
      cpg <- selected_cpg[[o]]
      df <- methylation_best_cpgs %>% filter(tile %in% cpg)
      if (o != "Multi_organ") df <- df %>% filter(Organ == o)
      df$organ_value <- o
      meth_list[[o]] <- df
    }
    merged_meth <- bind_rows(meth_list) %>%
      mutate(organ_value = factor(organ_value, levels = c("Brain","Caudal_Fin","Intestines","Multi_organ")),
             meth = methylation, CpG = tile)
  } else {
    df <- methylation_best_cpgs %>% filter(tile %in% selected_cpg)
    merged_meth <- df %>%
      mutate(Organ = factor(Organ, levels = c("Brain","Caudal_Fin","Intestines")),
             meth = methylation, CpG = tile)
  }

  n_cpgs_matched <- length(unique(merged_meth$CpG))

  if (n_cpgs_matched == 0) {
    cat("[", label, "] 0 of the selected CpGs are present in methylation_cpgs.csv",
        "(170 tiles total) - cannot compute drift direction for this set.\n\n")
    return(NULL)
  }

  slopes_all <- merged_meth %>%
    group_by(Organ, CpG) %>%
    summarise(
      slope     = coef(lm(meth ~ Age))[2],
      mean_meth = median(meth, na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    mutate(deviation = mean_meth - 0.5)

  set.seed(42)
  nboot <- 1000
  boot_cor <- slopes_all %>%
    group_by(Organ) %>%
    summarise(
      cor_boot = list({
        sapply(1:nboot, function(i) {
          samp <- sample(n(), replace = TRUE)
          cor(deviation[samp], slope[samp], use = "complete.obs")
        })
      }),
      .groups = "drop"
    )

  boot_summary <- boot_cor %>%
    rowwise() %>%
    mutate(
      mean_cor = mean(unlist(cor_boot)),
      se_cor   = sd(unlist(cor_boot)) / sqrt(length(unlist(cor_boot))),
      label    = label
    ) %>%
    ungroup()

  cat("[", label, "] CpGs matched:", n_cpgs_matched, "of", length(unlist(selected_cpg)), "selected\n")
  print(boot_summary %>% dplyr::select(Organ, mean_cor, se_cor))
  cat("\n")

  boot_summary
}

drift_tissue_specific <- run_drift_check(best_selected_cpgs_organs, "tissue-specific (Fig4D)", tissue_specific = TRUE)
drift_universal_all    <- run_drift_check(best_cpg_overlap_all, "universal, all 4 organs (Fig5D)")
drift_universal_single  <- run_drift_check(best_cpg_overlap_single, "universal, soma overlap (Brain;Fin;Intestines)")
```

```{r drift-plot}
all_drift <- bind_rows(drift_tissue_specific, drift_universal_all, drift_universal_single)

if (nrow(all_drift) > 0) {
  ggplot(all_drift, aes(x = Organ, y = mean_cor, fill = Organ)) +
    geom_col() +
    geom_errorbar(aes(ymin = mean_cor - se_cor, ymax = mean_cor + se_cor), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    facet_wrap(~label, scales = "free_x") +
    labs(y = "Correlation (deviation vs slope)", x = "",
         title = "Drift-direction re-check across CpG-set definitions") +
    theme_classic(base_size = 12) +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
}
```

# Interpretation

Part A shows directly whether Fig4D and Fig5D are duplicating the same
CpG evidence (a "one finding drawn twice" problem) or are genuinely
independent - compare the printed overlap counts against the "30 shared
loci" figure in the manuscript text.

Part B gives, per CpG-set definition, the sign and bootstrap CI of the
deviation-vs-slope correlation in each organ. Read the sign for each
organ/panel combination against Fig4D's caption and surrounding text; a
sign flip for the panel(s) actually underlying Fig4D is a direction
mismatch that needs resolving before submission.
