---
title: "Check: methylation entropy direction per organ"
author: "Jean-Charles de coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
SUPPLEMENTARY CHECK - not part of the main numbered pipeline (01-10).
=================================================================
Re-derives the two entropy measures behind Fig7C (CpG-wise Shannon
entropy and sample-wise/global entropy, per organ) so the direction of
the organ ranking can be read off directly and compared against Fig7C's
caption and surrounding text.

Reuses the entropy logic from 08_dynamic_network_pca_entropy.Rmd
verbatim, but is self-contained: it does not require 01-07 to have been
run first in the same session, and only needs files already tracked in
this repo (methylation_cpgs.csv + results_cpg_final.csv +
final_metadata_uniqueID_new.xlsx - none of these are in the excluded
large-file list, see .gitignore).
=================================================================
-->

```{r config}
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(readr)
  library(readxl)
  library(GenomicRanges)
  library(lme4)
  library(car)
  library(emmeans)
  library(ggplot2)
  library(ggbeeswarm)
})
```

# Load data (self-contained)

```{r load-data}
clean_up_data <- function(df) {
  df %>%
    filter(dataset == "Validation") %>%
    filter(!model %in% c("old_elastic_net", "old_lasso")) %>%
    mutate(
      rsq_threshold = dplyr::recode(rsq_threshold,
        "nofilter" = "r2_nofilter", "0.05" = "r2_0.05", "0.1" = "r2_0.1",
        "0.2" = "r2_0.2", "0.3" = "r2_0.3"),
      rsq_threshold = factor(rsq_threshold,
        levels = c("r2_nofilter","r2_0.05","r2_0.1","r2_0.2","r2_0.3")),
      filter = factor(filter, levels = c("nofilter","min5","min10","min20")),
      unique_model_id = paste(cpg_grouping, model_id, sep = "_")
    ) %>%
    mutate(organ = case_when(
      organ == "Fin_clip_caudal" ~ "Caudal_Fin",
      organ == "All_organs" ~ "Multi_organ",
      TRUE ~ organ
    ))
}

cpg_data <- read.csv(here("data/Metrics_all_models/final_merged/results_cpg_final.csv"))
cpg_data_validation <- clean_up_data(cpg_data)

cpg_data_validation_long <- cpg_data_validation %>%
  tidyr::separate_rows(selected_cpgs, sep = ";") %>%
  dplyr::mutate(selected_cpgs = trimws(selected_cpgs)) %>%
  mutate(selected_cpgs = gsub("chr", "", selected_cpgs)) %>%
  separate(selected_cpgs, c("chr", "start", "end"), remove = FALSE, sep = ":") %>%
  dplyr::mutate(chr = trimws(chr), chr = factor(chr, levels = c(1:25)))

metadata <- read_excel(here("data/final_metadata_uniqueID_new.xlsx"))

methylation_best_cpgs <- read_csv(here("data/Metrics_all_models/methylation_data/methylation_cpgs.csv"),
                                   show_col_types = FALSE) %>%
  mutate(
    Age = as.factor(Age),
    Organ = case_when(Organ == "Fin_clip_caudal" ~ "Caudal_Fin", TRUE ~ Organ)
  ) %>%
  left_join(metadata %>% dplyr::select(Sanger_new_ID, individual_fish_ID), by = "Sanger_new_ID")
```

# Rebuild the "universal, soma overlap" CpG set used by Fig7C

Verbatim reproduction of the overlap-computation logic from
`03_best_model_ranking_significance.Rmd`, restricted to what
`best_cpg_overlap_single` needs.

```{r overlap-computation}
fraction_threshold <- 0.05

cpg_ranges <- cpg_data_validation_long %>%
  dplyr::select(cpg_grouping, organ, selected_cpgs, chr, start, end) %>%
  mutate(selected_cpgs = trimws(selected_cpgs), start = as.numeric(start), end = as.numeric(end))

cpg_granges_list <- cpg_ranges %>%
  group_by(organ) %>%
  summarise(
    gr = list(GRanges(seqnames = chr, ranges = IRanges(start = start, end = end), cpg = selected_cpgs)),
    .groups = "drop"
  ) %>%
  deframe()

organ_names <- names(cpg_granges_list)
overlap_results <- list()

for (k in 2:length(organ_names)) {
  combs <- combn(organ_names, k, simplify = FALSE)
  overlap_results[[paste0("k", k)]] <- map_df(combs, function(orgs_combo) {
    gr_list <- cpg_granges_list[orgs_combo]
    cpg_vectors <- lapply(gr_list, function(gr) gr$cpg)
    intersect_cpgs <- Reduce(intersect, cpg_vectors)
    if (length(intersect_cpgs) == 0) return(NULL)
    tibble(organs = paste(orgs_combo, collapse = ";"), cpg = intersect_cpgs)
  })
}

k2_plus_df <- bind_rows(overlap_results, .id = "k")

best_cpg_overlap_single <- k2_plus_df %>%
  filter(organs == "Brain;Caudal_Fin;Intestines") %>%
  pull(cpg)

cat("CpGs in the universal soma-overlap set:", length(best_cpg_overlap_single), "\n")
```

# CpG-wise Shannon entropy per organ

Verbatim reproduction of the entropy logic in
`08_dynamic_network_pca_entropy.Rmd`.

```{r cpg-wise-entropy}
selected_cpg <- best_cpg_overlap_single

df <- methylation_best_cpgs %>% filter(tile %in% selected_cpg)

merged_meth_matrix <- df %>%
  dplyr::select(tile, methylation, Organ, individual_fish_ID)

tile_organ_means <- merged_meth_matrix %>%
  group_by(tile, Organ, individual_fish_ID) %>%
  summarise(methylation = mean(methylation, na.rm = TRUE), .groups = "drop")

entropy_per_cpg <- lapply(unique(tile_organ_means$Organ), function(org) {
  mat <- tile_organ_means %>% filter(Organ == org) %>%
    dplyr::select(tile, methylation, individual_fish_ID) %>%
    pivot_wider(names_from = individual_fish_ID, values_from = methylation)

  mat_numeric <- as.matrix(mat[, -1]) / 100
  mat_numeric <- pmin(pmax(mat_numeric, 1e-6), 1 - 1e-6)

  entropy <- apply(mat_numeric, 1, function(x) {
    mean(-x * log2(x) - (1 - x) * log2(1 - x))
  })

  data.frame(organ = org, tile = mat$tile, entropy = entropy)
}) %>% bind_rows()

p_cpg <- ggplot(entropy_per_cpg, aes(x = organ, y = entropy, colour = organ, fill = organ)) +
  geom_violin(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  stat_summary(fun = mean, geom = "point", color = "black", show.legend = FALSE) +
  stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width = 0.2, show.legend = FALSE) +
  labs(title = "CpG-wise methylation entropy per organ",
       y = "Methylation heterogeneity (0-1)", x = "", colour = "", fill = "") +
  ylim(0, 1) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_cpg

cat("\n=== CpG-wise entropy: mean per organ (ranked highest to lowest) ===\n")
entropy_per_cpg %>%
  group_by(organ) %>%
  summarise(mean_entropy = mean(entropy), median_entropy = median(entropy), n_cpgs = n()) %>%
  arrange(desc(mean_entropy)) %>%
  print()
```

# Sample-wise (global) entropy per organ

```{r sample-wise-entropy}
sample_entropy <- lapply(unique(tile_organ_means$Organ), function(org) {
  mat <- tile_organ_means %>% filter(Organ == org) %>%
    dplyr::select(tile, methylation, individual_fish_ID) %>%
    pivot_wider(names_from = tile, values_from = methylation)

  mat_numeric <- as.matrix(mat[, -1]) / 100
  mat_numeric <- pmin(pmax(mat_numeric, 1e-6), 1 - 1e-6)

  entropy <- apply(mat_numeric, 1, function(x) {
    mean(-x * log2(x) - (1 - x) * log2(1 - x))
  })

  data.frame(organ = org, sample_id = mat$individual_fish_ID, entropy = entropy)
}) %>% bind_rows()

p_sample <- ggplot(sample_entropy, aes(x = organ, y = entropy, colour = organ, fill = organ)) +
  geom_violin(alpha = 0.6) +
  ggbeeswarm::geom_quasirandom(alpha = 0.2) +
  stat_summary(fun = mean, geom = "point", color = "black", show.legend = FALSE) +
  stat_summary(fun.data = mean_se, geom = "errorbar", color = "black", width = 0.2, show.legend = FALSE) +
  labs(title = "Sample-wise methylation entropy per organ",
       y = "Methylation heterogeneity (0-1)", x = "", colour = "", fill = "") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 20, hjust = 1))

p_sample

cat("\n=== Sample-wise entropy: mean per organ (ranked highest to lowest) ===\n")
sample_entropy %>%
  group_by(organ) %>%
  summarise(mean_entropy = mean(entropy), median_entropy = median(entropy), n_samples = n()) %>%
  arrange(desc(mean_entropy)) %>%
  print()

m <- lmer(entropy ~ organ + (1 | sample_id), data = sample_entropy)
summary(m)
Anova(m, type = "III")
emmeans(m, pairwise ~ organ)
```

# Interpretation

Compare the two "ranked highest to lowest" tables above (and the
`emmeans` pairwise contrasts for the sample-wise model) against what
Fig7C's caption and the text discussing it currently claim about which
organ has the highest/lowest entropy. If the ranking here disagrees with
the manuscript statement, that confirms the figure-vs-text mismatch is
real and tells you which one needs to be corrected - the CpG-wise and
sample-wise measures are also worth checking against each other, since
they can legitimately disagree (heterogeneity across CpGs within a
sample vs. heterogeneity across samples for a given CpG are different
quantities).
