---
title: "Check: tissue-ageing-rate ranking on the exact 30-CpG overlapping panel"
author: "Jean-Charles de coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
SUPPLEMENTARY CHECK - not part of the main numbered pipeline (01-10).
=================================================================
The Delta-Age-based tissue-ageing-rate ranking (intestine fastest,
caudal fin slowest) was checked against a clock-free methylation-slope
test on the 170-CpG "robust CpG" panel (the panel used in
08_dynamic_network_pca_entropy.Rmd etc.), which instead showed brain
clearly the slowest and caudal fin/intestine statistically
indistinguishable (Kruskal-Wallis chi-sq = 11.9, df = 2, p = 0.003;
pairwise Wilcoxon brain-fin p = 0.005, brain-intestine p = 0.005,
fin-intestine p = 0.86). This re-runs that check on the exact 30-CpG
panel used elsewhere in the paper (best_cpg_overlap_single,
"Brain;Caudal_Fin;Intestines" - the panel behind Figure 4D/5D and 7C)
instead of the broader 170-CpG panel.

All 30 CpGs in best_cpg_overlap_single are a subset of the 170-CpG panel
in methylation_cpgs.csv (checked directly - 30 of 30 present), which is
tracked in this repo (only the raw per-organ .tsv matrices under
data/Metrics_all_models/methylation_data/ are excluded, not this file).
So this check is fully self-contained: it does not require 01-08 to have
been run first, and needs no excluded data.
=================================================================
-->

```{r config}
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(readr)
  library(GenomicRanges)
  library(ggplot2)
})
```

# Derive the exact 30-CpG panel

Verbatim reproduction of the overlap-computation logic from
`03_best_model_ranking_significance.Rmd`, restricted to what
`best_cpg_overlap_single` needs (the k=3 "Brain;Caudal_Fin;Intestines"
intersection).

```{r load-and-derive-panel}
clean_up_data <- function(df) {
  df %>%
    filter(dataset == "Validation") %>%
    filter(!model %in% c("old_elastic_net", "old_lasso")) %>%
    mutate(unique_model_id = paste(cpg_grouping, model_id, sep = "_")) %>%
    mutate(organ = case_when(
      organ == "Fin_clip_caudal" ~ "Caudal_Fin",
      organ == "All_organs" ~ "Multi_organ",
      TRUE ~ organ
    ))
}

cpg_data <- read.csv(here("data/Metrics_all_models/final_merged/results_cpg_final.csv"))
cpg_data_validation <- clean_up_data(cpg_data)

cpg_data_validation_long <- cpg_data_validation %>%
  tidyr::separate_rows(selected_cpgs, sep = ";") %>%
  dplyr::mutate(selected_cpgs = trimws(selected_cpgs)) %>%
  mutate(selected_cpgs = gsub("chr", "", selected_cpgs)) %>%
  separate(selected_cpgs, c("chr", "start", "end"), remove = FALSE, sep = ":")

cpg_ranges <- cpg_data_validation_long %>%
  dplyr::select(organ, selected_cpgs, chr, start, end) %>%
  mutate(start = as.numeric(start), end = as.numeric(end))

cpg_granges_list <- cpg_ranges %>%
  group_by(organ) %>%
  summarise(
    gr = list(GRanges(seqnames = chr, ranges = IRanges(start = start, end = end), cpg = selected_cpgs)),
    .groups = "drop"
  ) %>%
  deframe()

organ_names <- names(cpg_granges_list)
overlap_results <- list()
for (k in 2:length(organ_names)) {
  combs <- combn(organ_names, k, simplify = FALSE)
  overlap_results[[paste0("k", k)]] <- map_df(combs, function(orgs_combo) {
    gr_list <- cpg_granges_list[orgs_combo]
    cpg_vectors <- lapply(gr_list, function(gr) gr$cpg)
    intersect_cpgs <- Reduce(intersect, cpg_vectors)
    if (length(intersect_cpgs) == 0) return(NULL)
    tibble(organs = paste(orgs_combo, collapse = ";"), cpg = intersect_cpgs)
  })
}
k2_plus_df <- bind_rows(overlap_results, .id = "k")

best_cpg_overlap_single <- k2_plus_df %>%
  filter(organs == "Brain;Caudal_Fin;Intestines") %>%
  pull(cpg)

cat("Exact overlapping panel size:", length(best_cpg_overlap_single), "CpGs\n")
stopifnot(length(best_cpg_overlap_single) > 0)
```

# Load methylation values and check panel coverage

```{r load-methylation}
methylation_best_cpgs <- read_csv(here("data/Metrics_all_models/methylation_data/methylation_cpgs.csv"),
                                   show_col_types = FALSE) %>%
  mutate(Organ = case_when(Organ == "Fin_clip_caudal" ~ "Caudal_Fin", TRUE ~ Organ))

panel_170 <- unique(methylation_best_cpgs$tile)
matched_cpgs <- intersect(best_cpg_overlap_single, panel_170)

cat("Exact panel CpGs found in methylation_cpgs.csv:", length(matched_cpgs),
    "of", length(best_cpg_overlap_single), "\n")

if (length(matched_cpgs) < length(best_cpg_overlap_single)) {
  cat(length(best_cpg_overlap_single) - length(matched_cpgs),
      "CpGs from the exact panel are NOT in methylation_cpgs.csv and would need",
      "the excluded raw per-organ .tsv matrices to include.\n")
}
```

# Per-CpG, per-organ slope (exact panel only)

Reuses the same slope construction as
`07_methylation_drift_and_pcgt_enrichment.Rmd` (`lm(meth ~ Age)` per
CpG per organ), restricted to the exact panel and to the three organs
being ranked.

```{r slopes}
df_panel <- methylation_best_cpgs %>%
  filter(tile %in% matched_cpgs, Organ %in% c("Brain", "Caudal_Fin", "Intestines")) %>%
  mutate(Age_numeric = as.numeric(as.character(Age)))

slopes_exact_panel <- df_panel %>%
  group_by(Organ, tile) %>%
  summarise(slope = coef(lm(methylation ~ Age_numeric))[2], .groups = "drop")

cat("CpG x organ slope rows:", nrow(slopes_exact_panel), "\n")
table(slopes_exact_panel$Organ)
```

# Kruskal-Wallis + pairwise Wilcoxon on the exact panel

Same test structure as the original 170-CpG-panel check, for direct
comparison.

```{r kw-test}
kw <- kruskal.test(slope ~ Organ, data = slopes_exact_panel)
print(kw)

pw <- pairwise.wilcox.test(slopes_exact_panel$slope, slopes_exact_panel$Organ,
                            p.adjust.method = "none")
print(pw)

cat("\n=== Reference: 170-CpG panel result (already reported) ===\n")
cat("Kruskal-Wallis chi-sq = 11.9, df = 2, p = 0.003\n")
cat("Pairwise Wilcoxon: brain-fin p = 0.005, brain-intestine p = 0.005, fin-intestine p = 0.86\n")
```

```{r plot}
ggplot(slopes_exact_panel, aes(x = Organ, y = slope, colour = Organ, fill = Organ)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Methylation-vs-age slope per CpG, exact 30-CpG overlapping panel",
       y = "dM/dAge", x = "") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

# Interpretation

If the exact-panel result (Kruskal-Wallis p and the three pairwise
p-values above) agrees in direction with the 170-CpG-panel reference
numbers printed above - brain clearly slowest, fin and intestine not
distinguishable - that confirms the clock-free check is not an artefact
of which CpG panel was used, and strengthens the case for softening
"caudal fin ages slowest" in favour of "brain shows the most
stable/slowest methylation change." If the ranking instead changes on
this much smaller panel (30 vs 170 CpGs), treat that as a real
sensitivity to panel choice worth reporting explicitly rather than
picking whichever result is more convenient.
