---
title: "Check: model-count-matched subsampling robustness"
author: "Jean-Charles de coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
SUPPLEMENTARY CHECK - not part of the main numbered pipeline (01-05).
=================================================================
Point clocks have far fewer validation models per tissue (13-35) than
Cluster (309-348) or Tile (456-476) - see 01_setup_load_and_clean.Rmd /
02_best_model_ranking_significance.Rmd. This asks whether the "Tile >
Point/Cluster" ranking on performance_score survives once all three
groups are subsampled down to the same n.

Run 01_setup_load_and_clean.Rmd and 02_best_model_ranking_significance.Rmd
first in the same session (this script reuses `ranked` from there) - or
run the self-contained data-loading chunk below on its own.
=================================================================
-->

```{r config}
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(scales)
  library(lme4)
  library(car)
  library(ggplot2)
  library(purrr)
})
```

# Rebuild `ranked` (skip this chunk if already in your session)

```{r rebuild-ranked}
if (!exists("ranked")) {

  metrics_data <- read.csv(here("data/Metrics_all_models/final_merged/results_metrics_final.csv"))

  clean_up_data <- function(df) {
    df %>%
      filter(dataset == "Validation") %>%
      filter(!model %in% c("old_elastic_net", "old_lasso")) %>%
      mutate(
        rsq_threshold = dplyr::recode(rsq_threshold,
          "nofilter" = "r2_nofilter", "0.05" = "r2_0.05", "0.1" = "r2_0.1",
          "0.2" = "r2_0.2", "0.3" = "r2_0.3"),
        rsq_threshold = factor(rsq_threshold,
          levels = c("r2_nofilter","r2_0.05","r2_0.1","r2_0.2","r2_0.3")),
        filter = factor(filter, levels = c("nofilter","min5","min10","min20")),
        unique_model_id = paste(cpg_grouping, model_id, sep = "_")
      ) %>%
      mutate(organ = case_when(
        organ == "Fin_clip_caudal" ~ "Caudal_Fin",
        organ == "All_organs" ~ "Multi_organ",
        TRUE ~ organ
      ))
  }

  metrics_data_validation <- clean_up_data(metrics_data)

  ranked <- metrics_data_validation %>%
    group_by(cpg_grouping, model, organ) %>%
    mutate(
      mae_score = scales::rescale(-mae),
      r2_score  = scales::rescale(r_squared_value)
    ) %>%
    ungroup() %>%
    mutate(
      performance_score = mae_score + r2_score,
      cpg_grouping = factor(cpg_grouping),
      tile_methode = factor(tile_methode),
      model  = factor(model),
      organ  = factor(organ),
      epse_scaled = scale(epse)
    )
}

table(ranked$cpg_grouping)
```

# Full-data baseline (for comparison)

```{r baseline}
m_full <- lmer(
  performance_score ~ organ * cpg_grouping + (1 | model_id),
  data = ranked
)
Anova(m_full, type = "III")
```

# Subsampling robustness check

For each of `n_repeats` iterations: subsample Cluster and Tile down to the
same n as Point (matched **per tissue**, since the imbalance is present in
every tissue - see Figure 1A), refit the same organ x cpg_grouping model,
and record the CpG-grouping omnibus p-value and the three pairwise
estimates. If the "Tile > Point/Cluster" ranking is a real effect and not
an artefact of Point's small n, it should hold in most/all repeats.

```{r subsample-check}
set.seed(42)
n_repeats <- 200

point_n_per_tissue <- ranked %>%
  filter(cpg_grouping == "Point") %>%
  count(organ, name = "n_point")

subsample_once <- function(seed) {
  set.seed(seed)

  matched <- ranked %>%
    inner_join(point_n_per_tissue, by = "organ") %>%
    group_by(organ, cpg_grouping) %>%
    group_modify(function(df, key) {
      if (key$cpg_grouping == "Point") return(df)
      n_target <- unique(df$n_point)
      df %>% slice_sample(n = min(n_target, nrow(df)))
    }) %>%
    ungroup()

  m <- tryCatch(
    lmer(performance_score ~ organ * cpg_grouping + (1 | model_id), data = matched),
    error = function(e) NULL
  )
  if (is.null(m)) return(NULL)

  a <- tryCatch(Anova(m, type = "III"), error = function(e) NULL)
  if (is.null(a)) return(NULL)

  em <- tryCatch(
    emmeans::emmeans(m, pairwise ~ cpg_grouping)$contrasts %>% as.data.frame(),
    error = function(e) NULL
  )

  tibble(
    seed = seed,
    cpg_grouping_chisq = a["cpg_grouping", "Chisq"],
    cpg_grouping_p      = a["cpg_grouping", "Pr(>Chisq)"],
    cluster_tile_est = if (!is.null(em)) em$estimate[em$contrast == "Cluster - Tile"] else NA,
    cluster_tile_p   = if (!is.null(em)) em$p.value[em$contrast == "Cluster - Tile"] else NA,
    point_tile_est   = if (!is.null(em)) em$estimate[em$contrast == "Point - Tile"] else NA,
    point_tile_p     = if (!is.null(em)) em$p.value[em$contrast == "Point - Tile"] else NA
  )
}

results <- map_df(1:n_repeats, subsample_once)
```

# Summary

```{r summary}
cat("Repeats completed:", nrow(results), "of", n_repeats, "\n\n")

cat("=== CpG-grouping omnibus test across repeats ===\n")
cat("% of repeats significant (p < 0.05):",
    round(100 * mean(results$cpg_grouping_p < 0.05, na.rm = TRUE), 1), "%\n")
print(summary(results$cpg_grouping_chisq))

cat("\n=== Cluster vs Tile across repeats ===\n")
cat("% of repeats where Tile still wins (estimate < 0) and p < 0.05:",
    round(100 * mean(results$cluster_tile_est < 0 & results$cluster_tile_p < 0.05, na.rm = TRUE), 1), "%\n")

cat("\n=== Point vs Tile across repeats ===\n")
cat("% of repeats where Tile still wins (estimate < 0) and p < 0.05:",
    round(100 * mean(results$point_tile_est < 0 & results$point_tile_p < 0.05, na.rm = TRUE), 1), "%\n")
cat("Median estimate:", round(median(results$point_tile_est, na.rm = TRUE), 3),
    " | Median p:", round(median(results$point_tile_p, na.rm = TRUE), 3), "\n")
```

```{r plot}
p <- ggplot(results, aes(x = point_tile_est)) +
  geom_histogram(bins = 30, fill = "#C46E7A", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = paste0("Point vs Tile estimate across ", n_repeats, " count-matched subsamples"),
    x = "Estimate (negative = Tile outperforms Point)", y = "Count"
  ) +
  theme_classic(base_size = 12)

p
```

# Interpretation

If the histogram above sits consistently on one side of zero and most
repeats are significant, the "Tile > Point" (or "Tile > Cluster") finding
is unlikely to be a small-n artefact. If it straddles zero, or the
significant fraction is well below the ~95% you'd expect from a robust
effect, the original comparison is likely sensitive to Point's small
sample size, and this should be treated as a real limitation rather than
a caveat - tissue-stratified claims that hinge on Point specifically
would need softening.
