---
title: "Check: mean +/- CI performance across outer iterations"
author: "Jean-Charles de coriolis"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
=================================================================
SUPPLEMENTARY CHECK - not part of the main numbered pipeline (01-06).
=================================================================
Step 6 (02_clock_construction/step6_final_model_fitting/) keeps only the
single best-performing outer iteration out of 100 for the headline
result - with ~12 fish per held-out validation fold, this risks reported
performance being inflated relative to what a fresh dataset would show
(a winner's-curse pattern). The six step-6 scripts (point/cluster/tile x
elastic_net/lasso) record every outer iteration's held-out validation
metrics (val_r_squared, val_mae, n_cpg), not just the best one, so mean
+/- CI performance can be reported alongside the headline numbers. The
"keep best" logic that produces the headline numbers is unchanged, so
the single-best-iteration numbers already in results_metrics_final.csv
stay comparable.

This script needs two things run first, in order:

  1. Step 6, run on the HPC (the large raw per-organ methylation matrices
     it depends on are not part of this repo - see data/README.md).
  2. 03_combine_results/03_collate_outer_iteration_metrics.R, run once
     against the *_clock_modelled.rds outputs, producing
     results_outer_iterations.csv (long format: one row per model x
     outer iteration).

Point EPICLOCK_ROOT at a project root containing that CSV to run this as
a normal, self-contained analysis.
=================================================================
-->

```{r config}
PROJECT_ROOT <- Sys.getenv("EPICLOCK_ROOT", unset = normalizePath("."))
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(ggplot2)
})

outer_file <- here("results/results_outer_iterations.csv")
# If you ran 03_collate_outer_iteration_metrics.R with its default
# output_folder (under PROJECT_ROOT/multitissue/...), copy or symlink its
# results_outer_iterations.csv into this repo's results/ folder, or edit
# outer_file above to point at it directly.
```

```{r check-file-exists}
if (!file.exists(outer_file)) {
  stop(
    "results_outer_iterations.csv not found at: ", outer_file, "\n",
    "This means the two steps in the header comment above haven't been run ",
    "yet (run step 6, then run ",
    "03_combine_results/03_collate_outer_iteration_metrics.R). ",
    "This check cannot proceed without it - there is no way to recover ",
    "per-outer-iteration performance from .rds outputs that predate the ",
    "all_outer_validation_metrics field."
  )
}
```

# Load outer-iteration metrics and the existing single-best-iteration summary

```{r load-data}
outer_metrics <- read.csv(outer_file)

best_iteration_metrics <- read.csv(here("data/Metrics_all_models/final_merged/results_metrics_final.csv")) %>%
  filter(dataset == "Validation")

cat("Outer-iteration rows loaded:", nrow(outer_metrics), "\n")
cat("Unique model configurations (model_id):", n_distinct(outer_metrics$model_id), "\n")
cat("Outer iterations per model (should be up to 100):\n")
print(summary(as.numeric(table(outer_metrics$model_id))))
```

# Mean +/- 95% CI across outer iterations, per CpG-grouping x model config

```{r mean-ci-summary}
outer_summary <- outer_metrics %>%
  group_by(cpg_grouping, model, organ, rsq_threshold, filter) %>%
  summarise(
    n_outer      = n(),
    mean_r2      = mean(val_r_squared, na.rm = TRUE),
    sd_r2        = sd(val_r_squared, na.rm = TRUE),
    ci95_r2      = 1.96 * sd_r2 / sqrt(n_outer),
    mean_mae     = mean(val_mae, na.rm = TRUE),
    sd_mae       = sd(val_mae, na.rm = TRUE),
    ci95_mae     = 1.96 * sd_mae / sqrt(n_outer),
    .groups = "drop"
  )

outer_summary %>%
  arrange(desc(mean_r2)) %>%
  head(20) %>%
  print(width = Inf)
```

# Compare against the single-best-iteration numbers already reported

```{r compare-best-vs-mean}
comparison <- outer_summary %>%
  left_join(
    best_iteration_metrics %>%
      dplyr::select(cpg_grouping, model, organ, rsq_threshold, filter,
                    best_r2 = r_squared_value, best_mae = mae),
    by = c("cpg_grouping", "model", "organ", "rsq_threshold", "filter")
  ) %>%
  mutate(
    r2_inflation  = best_r2 - mean_r2,
    mae_deflation = mean_mae - best_mae
  )

cat("=== How much does picking the single best outer iteration inflate R^2",
    "relative to the 100-iteration mean? ===\n")
cat("Median R^2 inflation (best - mean):", round(median(comparison$r2_inflation, na.rm = TRUE), 3), "\n")
cat("Median MAE deflation (mean - best):", round(median(comparison$mae_deflation, na.rm = TRUE), 3), "\n\n")

cat("Fraction of configs where the reported best R^2 sits outside the",
    "95% CI of the 100-iteration mean:\n")
outside_ci <- comparison %>%
  mutate(outside = best_r2 > (mean_r2 + ci95_r2)) %>%
  summarise(pct = round(100 * mean(outside, na.rm = TRUE), 1))
print(outside_ci)
```

```{r plot}
ggplot(comparison, aes(x = mean_r2, y = best_r2, colour = cpg_grouping)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Reported best-iteration R^2 vs. 100-iteration mean R^2",
    x = "Mean R^2 across 100 outer iterations",
    y = "R^2 of the single reported (best) iteration",
    colour = ""
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

# Re-run the main CpG-grouping comparison using the mean instead of the best

```{r rerun-main-comparison}
suppressPackageStartupMessages({
  library(lme4)
  library(car)
  library(scales)
})

ranked_mean <- outer_summary %>%
  group_by(cpg_grouping, model, organ) %>%
  mutate(
    mae_score = scales::rescale(-mean_mae),
    r2_score  = scales::rescale(mean_r2)
  ) %>%
  ungroup() %>%
  mutate(performance_score = mae_score + r2_score)

m_mean <- lmer(
  performance_score ~ organ * cpg_grouping + (1 | rsq_threshold),
  data = ranked_mean
)
Anova(m_mean, type = "III")
```

# Interpretation

The plot above should sit mostly above the dashed y=x line if the
winner's-curse pattern is real - the single reported "best" iteration
should systematically look better than the honest 100-iteration mean.
The `outside_ci` figure quantifies how often the reported number is
outside what 100-iteration sampling variability alone would produce.

The final model refit (`ranked_mean`) repeats the main CpG-grouping
comparison using mean-across-outer-iterations performance instead of
best-of-100 - compare its `Anova()` output (chi-sq, p-values, direction
of effects) against the corresponding model fit on
`results_metrics_final.csv` in
`05_clock_design_benchmarking/02_best_model_ranking_significance.Rmd`
and `06_cross_tissue_ageing_biology/03_best_model_ranking_significance.Rmd`.
If the ranking/significance pattern holds, this is a reporting fix
(report mean +/- CI instead of best) with no knock-on effect on the
paper's conclusions. If it changes, the affected claims need revisiting.
